---
title: "Creating a model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating a model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE}
library(diseasy)
```

# Introduction
This vignette provides an overview of the `DiseasyModel` class and its use in creating a `diseasy` model.

As described in the package introduction (`vignette("diseasy")`), the power of `diseasy` comes from combining different
"functional" modules with model "templates" to create a large ensemble of models easily.

It is `DiseasyModel` that defines how such model templates should look, and provides the mechanism by which the
functional modules are loaded in the model templates to create the individual model members of the ensemble.

Before we dive into the details of `DiseasyModel`, let's first consider what it means to be a "model template":

# Model Templates
Developing a model in `diseasy` involves constructing a model template that incorporates one or more functional modules
into a model design.

For instance, let's consider the creation of a simple SIR (Susceptible, Infectious, Recovered) infection
model with a seasonal forcing $\sigma(t)$ of the overall infection rate, denoted by \beta.

The SEIR model with seasonal forcing can be represented by the following set of differential equations:

$$
\frac{dS(t)}{dt} = - \beta \cdot \sigma(t) \cdot S(t) \cdot I(t)
$$

$$
\frac{dI(t)}{dt} = \beta \cdot \sigma(t) \cdot S(t) \cdot I(t) - \gamma \cdot I(t)
$$

$$
\frac{dR(t)}{dt} = \gamma \cdot I(t)
$$

When developing this model traditionally, we would choose some functional form for the seasonal forcing $\sigma(t)$,
e.g. a sinusoidal relationship:
$$
\sigma(t) = A \cdot \cos(k \cdot t + \phi)
$$

The strategy of `diseasy` is to source the seasonal forcing $\sigma(t)$ directly from the corresponding functional module
(`DiseasySeason`).

Every implementation of a seasonal forcing in `DiseasySeason` has the same function signature
(i.e. they all take only a single argument $t$ once configured).

This way, we can configure different instances of `DiseasySeason` using the whole range of (available) plausible
functional forms of seasonal forcing:

* No seasonal forcing ($\sigma(t) = 1$)
* Sinusoidal forcing
* Temperature dependent forcing
* Other plausible seasonal forcing forms

Since the modules of `diseasy` are "objects" in an object-oriented-programming sense, the functional forms of the
seasonal forcing can have parameters (as we see in the list above), but these are stored in the object when configuring
the module, and the actual seasonal forcing exposed to the models all have the same function signature that takes
only time as input ($\sigma(t)$)[^1].

This way, we can easily swap out different seasonal forcing models in our model templates without having to change the
model equations themselves. If new seasonal forcing models are developed, they are then automatically available to the
developed model.

This is the philosophy behind the `diseasy` package, since we can easily swap out different functional modules for the
templates and generate a large ensemble of models with functional forms for the disease dynamics.

# The `DiseasyModel` class

To create a model template in `diseasy` is to create a new class that inherits the `DiseasyModel` class.

This gives the model template access to the functional modules that are available in the package, it defines
how the user sets the parameters of the model and queries the model for results.

In addition, it provides useful utilities for the implementation of the model such as logging and caching.

## A container for functional modules
The `DiseasyModel` class contains slots for instances of functional modules:

```{r generate_list_of_functional_modules, echo = FALSE, results = 'asis'}
# We dynamically update this list, as new functional modules are added to DiseasyModel
purrr::keep(names(DiseasyModel$private_fields), ~ startsWith(., ".Diseasy")) |>
  purrr::map_chr(~ stringr::str_remove(., stringr::fixed(".Diseasy"))) |>
  purrr::map_chr(~
    paste0(
      "- `$", stringr::str_to_lower(.), "`",
      ": `{Diseasy", ., "}`"
    )
  ) |>
  cat(sep = "\n")
```

This means, that once a new `DiseasyModel` instance is created, it will have the functional modules internally
available. Going back to the above example of the SIR model with seasonal forcing, the model template would be
access the seasonal forcing $\sigma(t)$ via the call `self$season$model_t(t)`.
```{r check_DiseasySeason_function_signature, include = FALSE}
# If the the signature of DiseasySeason is changed, please update this vignette
stopifnot("model_t" %in% names(DiseasySeason$active))
```

## Training data
The `DiseasyModel` class also contains the method `$get_training_data()` that provides training data to the models.

This method interfaces with the `DiseasyObservables` module to provide data for an observable at some stratification
level. The level of stratification is generally a flexible parameter that will be set by the user when requesting
results from the model. For example, the user my request results at the country level or at the regional level.

The helper method `$get_training_data()` will then provide the training data for the observable at the requested level.
This data will contain one record for each time point in the training period, with columns for the date of the
observation, the values of the different (optional) stratification, and the value of the observable on the date, and
finally, an integer counter for the day relative to the end of the training period.

## Model-user interface
Beyond providing access to the functional modules, `DiseasyModel` defines how the user will interface with the model.

### Setting parameters
The parameters of the model should be contained within the `$parameters` slot of the model template.

These are then exposed to the user, and logic can be implemented to allow the user to set some parameters of the model
while leaving others free to be determined by the model during model fitting.

#### Validating the parameters
The model template can test the user provided `parameters` by implementing the private `$validate_parameters()` method.
Note that some parameters may be set inherited from superclasses, so the implementation of the `$validate_parameters()`
method should also call the parent validation method (`super$validate_parameters()`).

### Getting results
Once the parameters of the model is set (either by the user or by the model), the user can query the model for results
using the `$get_results()` function. Through this function, the user provides the parameters of the result that should
be provided (target observable, time-window and stratification level). In return, the function returns a standardised
`data.frame`-like output that can be used for further analysis in conjunction with the results from other models.

The outputs of the model are standardised, and should follow the structure of the `training_data` provided by
`$get_training_data()` enriched by model prediction, model id and realization id.

[^1]: The seasonal model exposed by a configured `DiseasySeason` instance is a "closure" instead of a pure function.
