---
title: "Creating ensembles"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating ensembles}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE}
library(diseasy)
```

# Introduction

The goal of `diseasy` is to create and manipulate ensembles of disease models. This vignette will show how to create ensembles and how to use them to make predictions.


# Configuration of the models

Different models needs varying amount of configuration before they can enter into an ensemble.
To include a model template in your ensemble, you would need to consult the documentation for the template
to determine the steps needed to configure models from that template.

In this vignette, we create ensembles using the `DiseasyModelG0` and `DiseasyModelG1` templates which are
nearly configuration free.

## Observables module
All model template requires a configured observables module.

To configure this module we must:

- Specify the source of the data (i.e. which `diseasystore` to use).
  For this example, we will use the example data bundled with the package (`?DiseasystoreSeirExample`)
- Specify a data base connection.
  `diseasystore` requires us to connect to a data base to store features in.
  For this example, we will use an in-memory SQLite data base.
- Specify the `last_queryable_date` for the observables module.
  This date determine which data the models can "see", i.e. inform the ensemble, while everything
  after this date is kept from the ensemble.

```{r}
observables <- DiseasyObservables$new(
  diseasystore = DiseasystoreSeirExample,
  conn = DBI::dbConnect(RSQLite::SQLite())
)

observables$set_last_queryable_date(
  observables$ds$max_end_date - lubridate::days(50)
)
```


# Creating ensembles

With the functional modules configured, we can now use the `combineasy()` function to create an ensemble.

This function takes three arguments:

- *model_templates*: A list of model templates to use in the ensemble.
- *modules*: A combination of functional modules to load into each model template
- *parameters*: A combination of model parameters to load into each model template

All combinations of these three arguments are used to construct the ensemble, which means that the
size of the ensemble can rapidly increase.

## Example 1
To begin, we create a small ensemble using only one model template `DiseasyModelG1` and omitting the
optional season module. For the ensemble, we will supply three different parameter sets which changes
the training size of the model.

```{r}
ensemble <- combineasy(
  model_templates = list(DiseasyModelG1),
  modules = tidyr::expand_grid(
    observables = list(observables)
  ),
  parameters = tidyr::expand_grid(
    training_length = list(
      c("training" = 7,  "testing" = 0, "validation" = 0),
      c("training" = 14, "testing" = 0, "validation" = 0),
      c("training" = 21, "testing" = 0, "validation" = 0)
    )
  )
)
```
The output of `combineasy()` is simply a list of model instances which we give the class `DiseasyEnsemble`.

In doing so, we can use the `print()`, `summary()` and `plot()` functions to get a quick overview of the ensemble.

```{r}
print(ensemble)
```

```{r}
summary(ensemble)
```

To produce plots from the ensemble, we need to at least specify the following:
- *observable*: The observable to predict
- *prediction_length*: The number of days to predict

```{r}
plot(ensemble, observable = "n_positive", prediction_length = 30)
```

One of the advantages of `diseasy` is its ability to easily stratify the models at different levels.
In this example, the data in `DiseasystoreSeirExample` is stratified by age, and we can run our
ensemble on each age group separately:

```{r}
plot(
  ensemble,
  observable = "n_positive",
  stratification = rlang::quos(age_group),
  prediction_length = 30
)
```

These stratifications are *flexible* and we can stratify to our hearts desire, limited only by the
stratification level in the `diseasystore`:

```{r}
plot(
  ensemble,
  observable = "n_positive",
  stratification = rlang::quos(
    age_group = dplyr::case_match(
      age_group,
      "00-29" ~ "00-29",
      "30-59" ~ "30+",
      "60+" ~ "30+"
    )
  ),
  prediction_length = 30
)
```


The `plot()` method has additional optional arguments which can be used to customize the plot further:
- *by_model*: Should the results from the individual models be plotted instead?
- *context_length*: The number of observations leading up the prediction to include in the plot.

```{r}
plot(
  ensemble,
  observable = "n_positive",
  prediction_length = 30,
  context_length = 15,
  by_model = TRUE
)
```


## Example 2
Now we create a larger ensemble using both model template (`DiseasyModelG0` and `DiseasyModelG1`) and the parameter settings from before.

```{r}
ensemble <- combineasy(
  model_templates = list(DiseasyModelG0, DiseasyModelG1),
  modules = tidyr::expand_grid(
    observables = list(observables)
  ),
  parameters = tidyr::expand_grid(
    training_length = list(
      c("training" = 7,  "testing" = 0, "validation" = 0),
      c("training" = 14, "testing" = 0, "validation" = 0),
      c("training" = 21, "testing" = 0, "validation" = 0)
    )
  )
)
```

The ensemble now consists of $2 \times 3 \times 3 = 18$ models:
```{r}
print(ensemble)
```

```{r}
summary(ensemble)
```

And again we can use the `plot()` method to visualize the predictions from the ensemble:

```{r}
plot(ensemble, observable = "n_positive", prediction_length = 30)
```
