---
title: "DiseasyObservables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DiseasyObservables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(diseasy)
```

```{r vignette configuration, include = FALSE}
opts <- options(
  "tibble.print_min" = 5,
  "diseasy.logging" = FALSE
)

# We have a "hard" dependency for duckdb to render this vignette
suggests_available <- rlang::is_installed("duckdb")
```

# Introduction
The `DiseasyObservables` module is the module responsible for providing disease data to the models.
The module is primarily a wrapper around `diseasystores` which means the available data will depend on the specific
`diseasystore` being used.

# Configuring the module
The module needs some configuration to be initialized. Some of these can be set through options.
Primarily, we need to specify the `diseasystore`:

```{r initialization, eval = suggests_available}
obs <- DiseasyObservables$new(
  conn = DBI::dbConnect(duckdb::duckdb())
)
# NOTE: Alternatively we could set options("diseasy.conn" = ...)

obs$set_diseasystore(diseasystore = DiseasystoreSeirExample)
```

To see the data that comes with the underlying `diseasystore` we can query the module.
```{r available_observables, eval = suggests_available}
obs$available_observables
```

```{r available_stratifications, eval = suggests_available}
obs$available_stratifications
```

To check the current status of the module, the `$describe()` method can be used:
```{r describe, eval = suggests_available}
obs$describe()
```

# Getting observations
We can query the model to give the data for a given observable in a given time frame:
```{r example_1, eval = suggests_available}
obs$get_observation(
  observable = "n_positive",
  start_date = as.Date("2020-03-01"),
  end_date = as.Date("2020-03-03")
)
```

If we want to stratify our data, we supply the stratification argument.
This argument is designed to be flexible, but it means they need to be wrapped in `rlang::quos()`.
We will see below, why that is.
```{r example_2, eval = suggests_available}
obs$get_observation(
  observable = "n_admission",
  stratification = rlang::quos(age_group),
  start_date = as.Date("2020-03-01"),
  end_date = as.Date("2020-03-03")
) |>
  dplyr::arrange(date, age_group)
```

Since the stratification is flexible, we can programmatically stratify:
```{r example_3, eval = suggests_available}
obs$get_observation(
  observable = "n_admission",
  stratification = rlang::quos(
    young_age_groups = age_group == "00-29"
  ),
  start_date = as.Date("2020-03-01"),
  end_date = as.Date("2020-03-03")
) |>
  dplyr::arrange(date, young_age_groups)
```

# Creating "synthetic" observables
Sometimes you may want to customise the data provided by the observables module to you specific use case.

Within `?DiseasyObservables` we call such customised observables for "synthetic" observables in the sense that
you create (synthesize) the observables from existing features.

These observables can be created by using the `?DiseasyObservables$define_synthetic_observable()` method.

Let's consider some examples of when you may want to create a synthetic observable

## Renaming observables
You are working with a `diseasystore` where a observable of interest is named differently than a model expects.
Rather than modifying the `diseasystore` you can create a synthetic observable renames a existing observable:

```{r synthetic_observable_renaming_1, eval = suggests_available}
# The diseasystore uses "n_admission" but you prefer "n_new_to_hospital"
obs$define_synthetic_observable(
  name = "n_new_to_hospital",
  mapping = \(n_admission) n_admission
)

obs$get_observation(
  observable = "n_new_to_hospital",
  start_date = as.Date("2020-03-01"),
  end_date = as.Date("2020-03-03")
)
```

Note that the stratifications are still available for the synthetic observable:
```{r synthetic_observable_renaming_2, eval = suggests_available}
obs$get_observation(
  observable = "n_new_to_hospital",
  stratification = rlang::quos(
    young_age_groups = age_group == "00-29"
  ),
  start_date = as.Date("2020-03-01"),
  end_date = as.Date("2020-03-03")
) |>
  dplyr::arrange(date, young_age_groups)
```

## Adjusting existing observables
You have some observable, say the number of test-positive cases, but the model you are working with needs
the number of true infected.
You also have a estimate of the amount of under-reporting and want construct a adjusted
observable for the model to use.

```{r synthetic_observable_adjusting_1, eval = suggests_available}
obs$define_synthetic_observable(
  name = "n_true_infected",
  mapping = \(n_positive) round(n_positive / 0.65) # 65% of cases are detected
)

obs$get_observation(
  observable = "n_true_infected",
  start_date = as.Date("2020-03-01"),
  end_date = as.Date("2020-03-03")
)
```

## Constructing new observables from multiple observables
For this example, we want to add the incidence rate of a disease in the population.
This requires some signal for the number of infected in the population, for which we will use the
"n_true_infected" observable we created above. In addition, we need the population to calculate the rate.

```{r advanced_synthetic_observable_1, eval = suggests_available}
obs$define_synthetic_observable(
  name = "incidence",
  mapping = \(n_positive, n_population) n_positive / n_population
)

# First we see that it works with the stratification
obs$get_observation(
  observable = "incidence",
  stratification = rlang::quos(age_group),
  start_date = as.Date("2020-03-01"),
  end_date = as.Date("2020-03-03")
) |>
  dplyr::arrange(date, age_group)
```

```{r advanced_synthetic_observable_2, eval = suggests_available}
# And it works without stratification
obs$get_observation(
  observable = "incidence",
  start_date = as.Date("2020-03-01"),
  end_date = as.Date("2020-03-03")
)
```

```{r cleanup, include = FALSE}
if (exists("obs")) rm(obs)
gc()
options(opts)
```
