---
title: "DiseasyActivity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DiseasyActivity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(diseasy)
```

```{r options, include = FALSE}
withr::local_options("digits" = 2, "tibble.print_min" = 5)
```

# Introduction
The `DiseasyActivity` is the module responsible for the implementation of contact matrices and societal restrictions.

# Configuring the module
The module can be initialized without setting any parameters.
```{r initialization}
act <- DiseasyActivity$new()
```

By default, the module initializes with no activity scenario, and produces non-informative outputs.
(See [the interfacing section](#interfacing) for ways to get metrics from the module)
```{r null_scenario}
act$get_scenario_openness()
```

To use the module, we must define the activity scenario. In the example below, we configure the module to use
Danish activity units and the "contact basis" for Denmark (i.e basic contact matrices and population information).

NOTE: the given contact basis uses counts of contacts between individuals rather than rates.
Having contacts as counts is important if you want to project the contact matrices into different age groups.
(see `?contact_basis` for more details on how the `contact_basis` is constructed)
```{r dk_configuration}
act$set_activity_units(dk_activity_units)
act$set_contact_basis(contact_basis %.% DK)

# This configuration can also be done as part of the constructor
# act <- DiseasyActivity$new(activity_units = dk_activity_units
#                            contact_basis = contact_basis %.% DK)
```

# Defining an activity scenario
Once `activity_units` and `contact_basis` is supplied, we can configure the activity scenario.
In this example we have a single activity defined (the `baseline` activity) starting from 2023-01-01.
```{r configure_baseline}
act$change_activity(date = as.Date("2020-01-01"), opening = "baseline")
```

To get an overview of the loaded activities, we can use the `$get_scenario_activities()` method.
```{r configure_baseline_example}
act$get_scenario_activities()
```

Calls to `$change_activity` are cumulative, which means that we can expand our existing scenario:
```{r configure_lockdown}
act$change_activity(
  date = as.Date("2020-03-11"),
  opening = "lockdown_2020",
  closing = "baseline"
)
# NOTE: we also have to "close" the "baseline" activity unit.
# The activity described by this unit is no longer in effect
# while the "lockdown_2020" activity unit is active.

act$get_scenario_activities()
```

If we try to add the same `activity_unit` while it is already open, the module will give an error:
```{r configure_error, error = TRUE}
act$change_activity(date = as.Date("2020-03-11"), opening = "lockdown_2020")
```

By default, activity units are added with a risk modifier of 1. The module includes the option to set an overall risk
associated with each of the 4 types of activity ("home", "work", "school", "other").

Having the risk as a separate modifier, allows the activity units to describe the level of contacts between individuals
while the risk modifier can describe risk-mitigation measures that effect these contacts.

To modify the associated risk, we use the `$change_risk()` method.
```{r configure_risk}
# Mitigation measures in workplaces reduce the risk
act$change_risk(date = as.Date("2020-03-11"), type = "work", risk = 0.45)

act$risk_matrix
```

To create a new activity scenario with an active instance of the module, we can use the `$reset_scenario()` method:
```{r reset_scenario}
act$reset_scenario()
act$get_scenario_activities()
```


The scenario can also be loaded directly from a `data.frame` such as the `dk_reference_scenario`
```{r print_dk_reference}
print(dk_reference_scenario)
```

```{r configure_dk_reference}
act$change_activity(dk_reference_scenario)
activities <- act$get_scenario_activities()

# NOTE: The full list of activities is too long to print in this vignette
activities[1:3]
```

The loaded scenario can be "cropped" through the `$crop_scenario()` method.
```{r crop_scenario}
act$crop_scenario(as.Date("2020-03-01"), as.Date("2020-04-15"))
act$get_scenario_activities()

# NOTE: The full list of activities in the scenario is shown here
# NOTE: The start date of the "baseline" activity is now updated to the given date
```

# Interfacing with the activity scenario {#interfacing}

For these examples, we reduce the activity scenario to just a single activity:
```{r interfacing_reset_scenario}
act$reset_scenario()
act$change_activity(date = as.Date("2020-03-11"), opening = "baseline")
```

With the activity scenario loaded, we query the module in different ways to get contact metrics from the module for the
scenario.

## Contact matrices
Several disease models uses contact matrices to structure the populations.

To retrieve the contact matrices from the module, we use the `$get_scenario_contacts()` method:
```{r scenario_contacts}
# We can project the matrices into custom age-groups by providing the
# "age_cuts_lower" argument
act$get_scenario_contacts(age_cuts_lower = c(0, 60))
```
The `DiseasyActivity` module provides separate matrices for the four arenas: "home", "school", "work", and "other".
These matrices can be combined by supplying a `weights` argument to the `$get_scenario_contacts()` method.

```{r scenario_contacts_weighted}
act$get_scenario_contacts(
  age_cuts_lower = c(0, 20, 40, 60, 80),
  weights = c(1, 1, 1, 1)
)
```

## Societal openness
A simpler metric for societal activity is the degree of "openness", i.e. degree to which their activity is limited
relative to an open society.

This value ranges from 1 (normal activity) to 0 (no activity).
```{r scenario_openness}
act$get_scenario_openness()
```

The degree of openness can be queried similarly to the contact matrices:
```{r scenario_openness_weighted}
act$get_scenario_openness(
  age_cuts_lower = c(0, 20, 40, 60, 80),
  weights = c(1, 1, 1, 1)
)
```



# Inspecting the module
As we saw above, we can use the `print()` generic to get a human readable summary of the module
```{r inspect_describe}
print(act)
```

We can also inspect the individual elements of the module more carefully such as the `$scenario_matrix` and
`$risk_matrix`. Both provide the internal representation of the loaded scenario.
```{r inspect_scenario_matrix}
act$scenario_matrix
```

```{r inspect_risk_matrix}
act$risk_matrix
```
