---
title: "DiseasyObservables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DiseasyObservables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

options("tibble.print_min" = 5)
options("diseasy.logging" = FALSE)

# We have a "hard" dependency for duckdb to render this vignette
suggests_available <- rlang::is_installed("duckdb")
```

```{r setup}
library(diseasy)
```

# Introduction
The `DiseasyObservables` module is the module responsible for providing disease data to the models.
The module is primarily a wrapper around `diseasystores` which means the available data will depend on the specific
`diseasystore` being used.

# Configuring the module
The module needs some configuration to be initialized. Some of these can be set through options.
Primarily, we need to specify the `diseasystore`:

```{r initialization, eval = suggests_available}
obs <- DiseasyObservables$new(
  conn = DBI::dbConnect(duckdb::duckdb())
)
# NOTE: Alternatively we could set options("diseasy.conn" = ...)

obs$set_diseasystore(diseasystore = "Simulist")
```

To see the data that comes with the underlying `diseasystore` we can query the module.
```{r available_observables, eval = suggests_available}
obs$available_observables
```

```{r available_stratifications, eval = suggests_available}
obs$available_stratifications
```

To check the current status of the module, the `$describe()` method can be used:
```{r describe, eval = suggests_available}
obs$describe()
```

# Getting observations
We can query the model to give the data for a given observable in a given time frame:
```{r example_1, eval = suggests_available}
obs$get_observation(
  observable = "n_positive",
  start_date = as.Date("2020-03-01"),
  end_date = as.Date("2020-04-30")
)
```

If we want to stratify our data, we supply the stratification argument.
This argument is designed to be flexible, but it means they need to be wrapped in `rlang::quos()`.
We will see below, why that is.
```{r example_2, eval = suggests_available}
obs$get_observation(
  observable = "n_hospital",
  stratification = rlang::quos(age),
  start_date = as.Date("2020-03-01"),
  end_date = as.Date("2020-04-30")
)
```

Since the stratification is flexible, we can programmatically stratify:
```{r example_3, eval = suggests_available}
obs$get_observation(
  observable = "n_hospital",
  stratification = rlang::quos(
    young_age_groups = age < 20,
  ),
  start_date = as.Date("2020-03-01"),
  end_date = as.Date("2020-04-30")
)
```
