---
title: "DiseasyObservables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DiseasyObservables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

options("tibble.print_min" = 5)
```

```{r setup}
library(diseasy)
```

```{r download, include = FALSE}
# Download the first n rows of the Google data set of interest
tmp_dir <- stringr::str_replace_all(tempdir(), r"{\\}", .Platform$file.sep)
remote_conn <- options() %.% diseasystore.DiseasystoreGoogleCovid19.remote_conn
google_files <- c("by-age.csv", "demographics.csv", "index.csv", "weather.csv")

purrr::walk(google_files, \(file_name) {
  paste0(remote_conn, file_name) |>
    readr::read_csv(n_max = 1000, show_col_types = FALSE, progress = FALSE) |>
    readr::write_csv(file.path(tmp_dir, file_name))
})
options("diseasystore.DiseasystoreGoogleCovid19.source_conn" = tmp_dir)
options("diseasystore.DiseasystoreGoogleCovid19.verbose" = FALSE)
options("diseasy.logging" = FALSE)
```

# Introduction
The `DiseasyObservables` module is the module responsible for providing disease data to the models.
The module is primarily a wrapper around `diseasystores` which means the available data will depend on the specific
`diseasystore` being used.

# Configuring the module
The module needs some configuration to be initialized. Some of these can be set through options.
Primarily, we need to specify the `diseasystore`
```{r initialization}
obs <- DiseasyObservables$new(
  conn = DBI::dbConnect(RSQLite::SQLite())
)
# NOTE: Alternatively we could set options("diseasy.conn" = ...)

obs$set_case_definition(case_definition = "Google COVID-19")
```

To see the data that comes with underlying `diseasystore` we can query the module.
```{r available_observables}
obs$available_observables
```

```{r available_aggregations}
setdiff(obs$ds$available_features, obs$available_observables)
```

To check the current status of the module, the `$describe()` method can be used:
```{r describe}
obs$describe()
```

# Getting observations
We can query the model to gives the data for a given observable in a given time frame:
```{r example_1}
obs$get_observation(observable = "n_population",
                    start_date = as.Date("2020-03-01"),
                    end_date = as.Date("2020-05-01"))

# NOTE: "target_conn" is a SQLite data base which has "numeric" as the "date" type.
# NOTE: The population here is not stratified, so covers all countries in the Google data
```

If we want to stratify our data, we supply the aggregation argument.
This argument is designed to be flexible, but it means they need to be wrapped in `rlang::quos()`.
We will see below, why that is.
```{r example_2}
obs$get_observation(observable = "n_hospital",
                    aggregation = rlang::quos(age_group),
                    start_date = as.Date("2020-03-01"),
                    end_date = as.Date("2020-05-01"))
```

Since the aggregation is flexible, we can programmatically stratify:
```{r example_3}
obs$get_observation(observable = "n_hospital",
                    aggregation = rlang::quos(young_age_groups =
                                                age_group %in% c("00-09", "10-19")),
                    start_date = as.Date("2020-03-01"),
                    end_date = as.Date("2020-05-01"))

# NOTE: "target_conn" is a SQLite data base which has "numeric" as the "boolean" type.
```

