---
title: "Regressions models in `diseasy`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{diseasy-model-regression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(diseasy)
```

# Introduction
In `diseasy` we bundle some simple regression model templates and provide the tools to create custom regression
model templates.

# Overview
These are structured as R6 classes that uses inheritance to simplify the creating of individual templates.

Structurally, R6 classes we provide are as follows:

| Module               | Description        |
|--------------------- | ------------------ |
|`?DiseasyModel`       | All model templates initially inherit from this module (see `vignette("creating-a-model")`). |
|`?DiseasyRegression`  | Defines the structure of regression model templates (i.e. must implement `$fit_regression()`, `$get_prediction()`, and `$update_formula()`). |
|`?DiseasyModelGLM`    | Implements `$fit_regression()` and `$get_prediction()` using `stats::glm` |
|`?DiseasyModelBRM`    | Implements `$fit_regression()` and `$get_prediction()` using `brms::brm`  |
|`?DiseasyModelG0`     | Implements `$update_formula()` to create a constant predictor model (from `?DiseasyModelGLM`) |
|`?DiseasyModelG1`     | Implements `$update_formula()` to create a exponential growth predictor model (from `?DiseasyModelGLM`) |
|`?DiseasyModelB0`     | Implements `$update_formula()` to create a constant predictor model (from `?DiseasyModelBRM`) |
|`?DiseasyModelB1`     | Implements `$update_formula()` to create a exponential growth predictor model (from `?DiseasyModelBRM`) |

# Examples
Here we show how to use the simple regression model templates in `diseasy`.

The regression models are "duplicated" in the sense that there exist two similar
models implemented in different statistical engines. The predictions from the
models are very similar.

| Engine \ model | constant         | exponential      |
|----------------|------------------|------------------|
| *glm*          | `DiseasyModelG0` | `DiseasyModelG1` |
| *brms*         | `DiseasyModelB0` | `DiseasyModelB1` |


## The data to model
As our example data, we will use the bundled example data `?DiseasystoreSeirExample`.

```{r observables, eval = rlang::is_installed("RSQLite")}
# Configure a observables module with the example data and database
observables <- DiseasyObservables$new(
  diseasystore = DiseasystoreSeirExample,
  conn = DBI::dbConnect(RSQLite::SQLite())
)
```

The data is stratified by age and covers a single infection wave.
It contains two observables of interest for our example:

- *n_positive*: The observed subset of infected
  (65% of true infected with weekday variation in detection).
- *n_admissions*: The number of new hospital admissions due to infections
  (risk of hospitalisation dependent on age group).

When running the models, we should also define at what point in time they should provide predictions for.
This is done by setting the `last_queryable_date` in the observables module:

```{r last_queryable_date, eval = rlang::is_installed("RSQLite")}
last_queryable_date <- observables$ds$min_start_date + 40
```

Lets first look at the data we have available:
```{r example data positive, eval = rlang::is_installed(c("ggplot2", "RSQLite")), fig.alt = "Data to model: Number of positive cases."}
observables$get_observation(
  observable = "n_positive",
  stratification = rlang::quos(age_group),
  start_date = observables$ds$min_start_date,
  end_date = observables$ds$max_end_date
) |>
  ggplot2::ggplot(ggplot2::aes(x = date, y = n_positive)) +
  ggplot2::geom_vline(
    xintercept = last_queryable_date,
    color = "black",
    linetype = "dashed"
  ) +
  ggplot2::geom_point(color = "seagreen") +
  ggplot2::labs(title = "Observed number of positive cases") +
  ggplot2::facet_grid(~ age_group)
```

```{r example data admission, eval = rlang::is_installed(c("ggplot2", "RSQLite")), fig.alt = "Data to model: Number of hospital admissions."}
observables$get_observation(
  observable = "n_admission",
  stratification = rlang::quos(age_group),
  start_date = observables$ds$min_start_date,
  end_date = observables$ds$max_end_date
) |>
  ggplot2::ggplot(ggplot2::aes(x = date, y = n_admission)) +
  ggplot2::geom_vline(
    xintercept = last_queryable_date,
    color = "black",
    linetype = "dashed"
  ) +
  ggplot2::geom_point(color = "violetred3") +
  ggplot2::labs(title = "Number of admissions") +
  ggplot2::facet_grid(~ age_group)
```

## Applying the models to the data
The regressions models are special class of models which can directly model the data,
and requires essentially no configuration.

Since the models are "duplicated" as described above, we here show only the results
for the GLM family of models

```{r}
observables$set_last_queryable_date(last_queryable_date)
model_1 <- DiseasyModelG0$new(observables = observables)
model_2 <- DiseasyModelG1$new(observables = observables)
```

### Example plots for the constant predictor model
```{r}
plot(
  model_1,
  observable = "n_positive",
  prediction_length = 30
)
```

```{r}
plot(
  model_1,
  observable = "n_positive",
  stratification = rlang::quos(age_group),
  prediction_length = 30
)
```

### Example plots for the exponential growth predictor model
```{r}
plot(
  model_2,
  observable = "n_admission",
  prediction_length = 30
)
```

```{r}
plot(
  model_2,
  observable = "n_admission",
  stratification = rlang::quos(age_group),
  prediction_length = 30
)
```

```{r cleanup, include = FALSE}
if (exists("model_1")) rm(model_1)
if (exists("model_2")) rm(model_2)
gc()
```
