---
title: "DiseasySeason"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DiseasySeason}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(diseasy)
```

# Introduction
The `DiseasySeason` is the module responsible for the implementation of different models for the seasonal dependency of the disease.

# Configuring the module
The module can be initialized without setting any parameters. Note that some models require the parameter `reference_date` to be set which can be done during initialization of the module.
```{r initialization}
s <- DiseasySeason$new(reference_date = as.Date("2020-03-01"))
```
To see the season models that comes with the module we can query the module.
```{r available_models}
s$available_season_models
```

By default, the module initializes with the `constant_season()` model.
To check the current status of the module, including which model is set, the `$describe()` method can be used:
```{r describe}
s$describe()
```

Once a season model is set, interfacing with the model happens through the `$model_t()` and `$model_date()` methods.
```{r plot_constant_season_t}
t <- 0:365
plot(t, purrr::map_dbl(t, s$model_t),
     type = "l", lwd = 2,
     xlab = "Days past reference date",
     ylab = "Relative effect of season",
     ylim = c(0, 1.25), xaxs = 'i', yaxs = 'i')
```

```{r plot_constant_season_date}
d <- s$reference_date + 0:365
plot(d, purrr::map_dbl(d, s$model_date),
     type = "l", lwd = 2,
     xlab = "",
     ylab = "Relative effect of season",
     ylim = c(0, 1.25), xaxs = 'i', yaxs = 'i')
```

To use the other models for season, the module provides functions of the form `$use_*_season()`.
For example, to configure the module to use a cosine season instead we can use `$use_cosine_season()`
```{r configure_cosine_season}
s$use_cosine_season()
```

Once re-configured the models stored in `$model_t` and `$model_date` change to the new model:
```{r plot_cosine_season_date}
d <- s$reference_date + 0:365
plot(d, purrr::map_dbl(d, s$model_date),
     type = "l", lwd = 2,
     xlab = "",
     ylab = "Relative effect of season",
     ylim = c(0, 1.25), xaxs = 'i', yaxs = 'i')
```

Note that all models, other than `constant_season` has a `scale` parameter.
The scale parameter sets the relative difference between minimum and maximum season effect in the model.
For example, if `scale = 0.25`, the effect of season should never drop below 25% of the maximum.

```{r plot_cosine_season_date_w_scale}
s$set_scale(0.25)

d <- s$reference_date + 0:365
plot(d, purrr::map_dbl(d, s$model_date),
     type = "l", lwd = 2,
     xlab = "",
     ylab = "Relative effect of season",
     ylim = c(0, 1.25), xaxs = 'i', yaxs = 'i')

```

The models may have additional parameters that control their shape. See the documentation of the module to learn more about the specific models.
