---
title: "diseasy: quick start guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{diseasy: quick start guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(diseasy)
```

```{r hidden_options, include = FALSE}
if (rlang::is_installed("withr")) {
  withr::local_options("diseasystore.verbose" = FALSE)
} else {
  opts <- options("diseasystore.verbose" = FALSE)
}
```

# Overview
The `diseasy` package is large in scope as it aims to provide the infrastructure to easily create a large ensemble of
disease models.

This building of disease models is achieved via a modular design philosophy. To that end, we have a number of
"functional" modules that feed into "model" modules.

## Functional modules
The functional modules are smaller modules meant for defined purposes. These modules are "atomic" and form the building
blocks of the ensembles.

The currently available functional modules are:

```{r vignette_string, echo = FALSE}
functional_modules <- ls("package:diseasy") |>
  purrr::keep(~ stringr::str_detect(., pattern = "Diseasy(?!(Base|Model))"))
vignette_string <- paste0("\`vignette(\"", functional_modules, "\")\`", collapse = ", ")
```

```{r available_functional_modules}
ls("package:diseasy") |>
  purrr::keep(~ stringr::str_detect(., pattern = "Diseasy(?!(BaseModule|Model.*))"))
```

Each functional module is restricted to covering a single functionality for use in the models
(and sometimes for other functional modules).

For example, `DiseasySeason` defines a number of models for the season dependency of the disease.
This module has a single defined purpose (modelling the effect of season) but it provides a number of different models
of describing possible season dependencies:

```{r available_season_models}
s <- DiseasySeason$new()
s$available_season_models
```

As we will see later, these different models for season dependency will be used to construct the ensembles.

Each of these season models have similar functional signatures (i.e. the function calls to the models are similar).
This way, when we create a disease model, we can use the module to implement different models for season dependency in
the disease models easily.

The other modules have other functionality they implement.
For example, the module `DiseasyObservables` is responsible for providing data to the models and modules.
For `DiseasySeason`, this means that some models of season dependency can use real-world data, such as temperature, in
the season models.


The use of these modules is varied and each module comes with their own vignette describing their usage:
`r vignette_string`.


## Model modules
As stated above, the functional modules provide functionality for the disease models.

The currently available model modules are:
```{r available_model_modules}
ls("package:diseasy", pattern = "DiseasyModel.+")
```

Presently, no models are included in the package. These will come in the following releases.

```{r models_now_available, include = FALSE}
# If the model are now included, please continue the writing of this vignette
stopifnot(identical(ls("package:diseasy", pattern = "DiseasyModel.+"), character(0)))
```


## Configuring
All `diseasy` modules are build on top of the `DiseasyBaseModule` class.
This class provides a number of features that are common to all modules, such as logging and caching.

### Caching
Internally, the modules may use `{cachem}` to cache computations either in memory or on disk.
To control how this caching is done, you can provide the `cache` argument to the module constructor containing a
`cachem` cache object, or you can set the option `diseasy.cache` to a `cachem` cache object.
See `?DiseasyBaseModule` for more information.


```{r cleanup, include = FALSE}
rm(s)
gc()
if (!rlang::is_installed("withr")) {
  options(opts)
}
```
