---
title: "Creating ensembles"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating ensembles}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE}
library(diseasy)
```

# Introduction

The goal of `diseasy` is to create and manipulate ensembles of disease models. This vignette will show how to create ensembles and how to use them to make predictions.


# Configuration of the models

Different models needs varying amount of configuration before they can enter into an ensemble.
To include a model template in your ensemble, you would need to consult the documentation for the template
to determine the steps needed to configure models from that template.

In this vignette, we create ensembles using the `DiseasyModelG0` and `DiseasyModelG1` templates which are
nearly configuration free.

## Observables module
All model template requires a configured observables module.

To configure this module we must:

- Specify the source of the data (i.e. which `diseasystore` to use).
  For this example, we will use the example data bundled with the package (`?DiseasystoreSeirExample`)
- Specify a data base connection.
  `diseasystore` requires us to connect to a data base to store features in.
  For this example, we will use an in-memory SQLite data base.
- Specify the `last_queryable_date` for the observables module.
  This date determine which data the models can "see", i.e. inform the ensemble, while everything
  after this date is kept from the ensemble.

```{r, eval = rlang::is_installed("RSQLite")}
observables <- DiseasyObservables$new(
  diseasystore = DiseasystoreSeirExample,
  conn = DBI::dbConnect(RSQLite::SQLite())
)

observables$set_last_queryable_date(
  observables$ds$max_end_date - lubridate::days(50)
)
```


# Creating ensembles

With the functional modules configured, we can now use the `combineasy()` function to create an ensemble.

This function takes three arguments:

- *model_templates*: A list of model templates to use in the ensemble.
- *modules*: A combination of functional modules to load into each model template
- *parameters*: A combination of model parameters to load into each model template

All combinations of these three arguments are used to construct the ensemble, which means that the
size of the ensemble can rapidly increase.

## Example 1
To begin, we create a small ensemble using only one model template `DiseasyModelG1` and omitting the
optional season module. For the ensemble, we will supply three different parameter sets which changes
the training size of the model.

```{r, eval = rlang::is_installed("RSQLite")}
ensemble <- combineasy(
  model_templates = list(DiseasyModelG1),
  modules = tidyr::expand_grid(
    observables = list(observables)
  ),
  parameters = tidyr::expand_grid(
    training_length = list(
      c("training" = 7,  "testing" = 0, "validation" = 0),
      c("training" = 14, "testing" = 0, "validation" = 0),
      c("training" = 21, "testing" = 0, "validation" = 0)
    )
  )
)
```
The output of `combineasy()` is simply a list of model instances which we give the class `DiseasyEnsemble`.

In doing so, we can use the `print()`, `summary()`, `predict()` and `plot()` functions to get a quick overview
of the ensemble.

```{r, eval = rlang::is_installed("RSQLite")}
print(ensemble)
```

```{r, eval = rlang::is_installed("RSQLite")}
summary(ensemble)
```

To produce predictions and plots from the ensemble, we need to at least specify the following:

- *observable*: The observable to predict
- *prediction_length*: The number of days to predict

```{r, eval = rlang::is_installed("RSQLite")}
predict(ensemble, observable = "n_positive", prediction_length = 30) |>
  head()
```

```{r, fig.alt = "Plot of the ensemble predictions. Shaded area is ensemble quantiles with observations as points.", eval = rlang::is_installed("RSQLite")}
plot(ensemble, observable = "n_positive", prediction_length = 30)
```

One of the advantages of `diseasy` is its ability to easily stratify the models at different levels.
In this example, the data in `DiseasystoreSeirExample` is stratified by age, and we can run our
ensemble on each age group separately:

```{r, fig.alt = "Plot of the ensemble predictions stratified by age group. Shaded area is ensemble quantiles with observations as points.", eval = rlang::is_installed("RSQLite")}
plot(
  ensemble,
  observable = "n_positive",
  stratification = rlang::quos(age_group),
  prediction_length = 30
)
```

These stratifications are *flexible* and we can stratify to our hearts desire, limited only by the
stratification level in the `diseasystore`:

```{r, fig.alt = "Plot of the ensemble predictions stratified by custom age group. Shaded area is ensemble quantiles with observations as points.", eval = rlang::is_installed("RSQLite")}
plot(
  ensemble,
  observable = "n_positive",
  stratification = rlang::quos(
    age_group = dplyr::case_match(
      age_group,
      "00-29" ~ "00-29",
      "30-59" ~ "30+",
      "60+" ~ "30+"
    )
  ),
  prediction_length = 30
)
```


The `plot()` method has additional optional arguments which can be used to customize the plot further:

- *by_model*: Should the results from the individual models be plotted instead?
- *context_length*: The number of observations leading up the prediction to include in the plot.

```{r, fig.alt = "Plot of the ensemble predictions my model. Shaded area is ensemble quantiles with observations as points.", eval = rlang::is_installed("RSQLite")}
plot(
  ensemble,
  observable = "n_positive",
  prediction_length = 30,
  context_length = 15,
  by_model = TRUE
)
```


## Example 2
Now we create a larger ensemble using both model template (`DiseasyModelG0` and `DiseasyModelG1`) and the parameter settings from before.

```{r, eval = rlang::is_installed("RSQLite")}
ensemble <- combineasy(
  model_templates = list(DiseasyModelG0, DiseasyModelG1),
  modules = tidyr::expand_grid(
    observables = list(observables)
  ),
  parameters = tidyr::expand_grid(
    training_length = list(
      c("training" = 7,  "testing" = 0, "validation" = 0),
      c("training" = 14, "testing" = 0, "validation" = 0),
      c("training" = 21, "testing" = 0, "validation" = 0)
    )
  )
)
```

The ensemble now consists of $2 \times 3 = 6$ models:
```{r, eval = rlang::is_installed("RSQLite")}
print(ensemble)
```

```{r, eval = rlang::is_installed("RSQLite")}
summary(ensemble)
```

And again we can use the `plot()` method to visualize the predictions from the ensemble:

```{r, fig.alt = "Plot of predictions from alternative ensemble. Shaded area is ensemble quantiles with observations as points.", eval = rlang::is_installed("RSQLite")}
plot(ensemble, observable = "n_positive", prediction_length = 30)
```

## Example 3 - A large SEIR model ensemble`

In this example, we will create a large ensemble of SEIR models using the `DiseasyModelOdeSeir` template.

To do so, we will create a few different instances of the functional modules, `DiseasyActivity`, `DiseasySeason` and
`DiseasyImmunity`. Which each represent different mechanistic scenarios to include in the model.

### Activity module
```{r, eval = FALSE}
# Create three different contact bases for the ensemble
# All uses the Danish populations but with different contact matrices
contact_basis_1 <- contact_basis_2 <- contact_basis_3 <- contact_basis$DK
contact_basis_2$contacts <- contact_basis$SE$contact
contact_basis_3$contacts <- contact_basis$NO$contact

# Create an activity instance with the Danish changes in restrictions
activity <- DiseasyActivity$new()
activity$set_activity_units(dk_activity_units)
activity$change_activity(date = as.Date("2020-01-01"), opening = "baseline")

# Load the different contact bases to create distinct activity instances
activities <- list(
  contact_basis_1,
  contact_basis_2,
  contact_basis_3
) |>
  purrr::map(~ {
    act <- activity$clone()
    act$set_contact_basis(contact_basis = .)
    return(act)
  })
rm(activity)
```

### Season module
```{r, eval = FALSE}
# Create three different season modules for the ensemble
season <- DiseasySeason$new(reference_date = as.Date("2020-01-01"))

seasons <- c("constant_season", "cosine_season", "covid_season_v1") |>
  purrr::map(~ {
    s <- season$clone()
    s$use_season_model(.)
    return(s)
  })
rm(season)
```

### Immunity module
```{r, eval = FALSE}
# Use different waning immunity models for the ensemble
no_immunity <- DiseasyImmunity$new()
no_immunity$set_waning_model("no_waning")

# Exponential waning
immunities <- c(60, 120, 180) |>
  purrr::map(~ {
    immunity <- DiseasyImmunity$new()
    immunity$set_exponential_waning(time_scale = .)
    return(immunity)
  }) |>
  c(no_immunity)
```


### Observables module
Beyond setting the different scenarios for the models, we also need to define the "incidence" observable for the
`DiseasyModelOdeSeir` class.
We here use the number of observed positive cases adjusted for under-reporting.
```{r, eval = FALSE}
# Create an estimator for the incidence
observables$define_synthetic_observable(
  name = "incidence",
  mapping = \(n_positive, n_population) n_positive / (n_population * 0.65)
)
```

### Create the ensemble
With the functional modules configured, we can now use the `combineasy()` function to create a large SEIR ensemble.
We create the models with different compartment structures
(i.e. different number of exposed, infected and recovered compartments) to provide structurally different models.
```{r, eval = FALSE}
# Create the ensemble
seir_ensemble <- combineasy(
  model_templates = list(DiseasyModelOdeSeir),
  modules = tidyr::expand_grid(
    activity = activities,
    season = seasons,
    immunity = immunities,
    observables = list(observables)
  ),
  parameters = tidyr::expand_grid(
    # Hyper parameters
    "compartment_structure" = list(
      c("E" = 1L, "I" = 1L, "R" = 1L),
      c("E" = 2L, "I" = 1L, "R" = 1L),
      c("E" = 2L, "I" = 3L, "R" = 1L),
      c("E" = 1L, "I" = 1L, "R" = 3L),
      c("E" = 2L, "I" = 1L, "R" = 3L),
      c("E" = 2L, "I" = 3L, "R" = 3L)
    ),
    "age_cuts_lower" = list(
      c(0, 30, 60)
    ),
    # Parameters
    "disease_progression_rates" = list(
      c("E" = 1 / 2.1, "I" = 1 / 4.5)
    ),
    "overall_infection_risk" = list(
      0.025
    )
  )
)

summary(seir_ensemble)
```

```{r, fig.alt = "Plot of predictions from SEIR ensemble. Shaded area is ensemble quantiles with observations as points.", eval = FALSE}
plot(seir_ensemble, observable = "incidence", prediction_length = 30)
```
