---
title: "DiseasyRegions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DiseasyRegions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(diseasy)
```

```{r hidden_options, include = FALSE}
withr::local_options("diseasystore.verbose" = FALSE)
withr::local_options("tibble.print_min" = 5)
```

# Overview

The `DiseasyRegions` module defines the regional scope for model models.
In addition, it provides a common place to define the demography available for those regions,
and the adjacency between regions.

The module is intentionally generic. Region identifiers are treated as ordinary character values and are matched
exactly. This means that the region `"north"` is not interpreted as a parent of `"north_subregion"`.

# Data structure

`DiseasyRegions` uses three inputs:

- `regions`: a `character()` vector defining the selected region scope.
- `demography`: a `data.frame` with one region column, optional stratification columns, and one population column.
- `adjacency`: a `data.frame` describing connectedness between pairs of regions.

The demography data must include the columns `region` and `population`. Any additional columns are treated as
stratification variables.

```{r demography_data}
all_regions <- c("north", "south", "east", "north_subregion")

demography <- data.frame(
  region = rep(all_regions, each = 2),
  age_group = rep(c("0-17", "18+"), times = length(all_regions)),
  population = c(
    100000, 250000,
    80000, 180000,
    60000, 140000,
    20000, 50000
  )
)

demography
```

The adjacency data must include exactly the columns `from`, `to`, and `adjacency`.
The `from` and `to` columns should contain the same set of region identifiers.

```{r adjacency_data}
adjacency <- tidyr::expand_grid(
  from = all_regions,
  to = all_regions
)

adjacency[["adjacency"]] <- 0.1
adjacency[adjacency$from == adjacency$to, "adjacency"] <- 1
adjacency[adjacency$from == "north" & adjacency$to == "south", "adjacency"] <- 0.5
adjacency[adjacency$from == "south" & adjacency$to == "north", "adjacency"] <- 0.5

adjacency
```

# Creating a regional module

A `DiseasyRegions` object can be created and then configured with the selected regions, demography, and adjacency
data. Each setter validates the candidate input before storing it.

```{r create_module}
region <- DiseasyRegions$new()
region$set_regions(regions = c("north", "south", "east"))
region$set_demography(demography = demography)
region$set_adjacency(adjacency = adjacency)

region
```

The `?DiseasyRegions$demography` field returns demography filtered to the currently selected regions.

```{r filtered_demography}
region %.% demography
```

The active bindings `?DiseasyRegions$adjacency` and `?DiseasyRegions$adjacency_matirx`
expose derived views of this input after filtering to the selected regions.

```{r filtered_adjacency}
region %.% adjacency
```

# Adjacency matrix binding

The stored adjacency input is long form, but the module also exposes an `?DiseasyRegions$adjacency_matirx` field.
This field returns a matrix representation of adjacency for the currently selected regions.

The matrix is derived in three steps:

- the stored adjacency data are filtered to the selected regions;
- the filtered adjacency is made symmetric;
- the symmetric matrix is normalised so that row sums and column sums are both one, up to numerical tolerance.

```{r adjacency_matrix}
region %.% adjacency_matrix
```

The `adjacency` active binding is unpacked from this matrix. It returns the upper triangle of the normalised
symmetric matrix, including the diagonal.

```{r adjacency_from_matrix}
region %.% adjacency
```

This means that `region %.% adjacency` is not necessarily identical to the raw adjacency data supplied to
`$set_adjacency()`. It is a selected, symmetric, normalised view of those data.

Population totals can be computed directly from the filtered demography.

```{r population_totals}
aggregate(
  population ~ region,
  data = region %.% demography,
  FUN = sum
)
```

# Changing the selected regions

The selected regional scope can be changed with `$set_regions()`. The new regions must be available in the stored
adjacency and demography data.

```{r change_regions}
region$set_regions(regions = c("north", "south"))

region %.% regions
```

The active fields now reflect the updated scope.

```{r changed_demography}
region %.% demography
```

```{r changed_adjacency}
region %.% adjacency
```

```{r changed_adjacency_matrix}
region %.% adjacency_matrix
```

# Exact matching

The generic module does not use hierarchical matching. In the example data, `"north"` and `"north_subregion"` are two
different region identifiers.

```{r exact_matching}
region$set_regions(regions = "north")

region %.% demography
```

The result includes rows for `"north"`, but not for `"north_subregion"`.

# Updating demography and adjacency

The demography and adjacency can also be updated after the module has been created. Candidate data are validated before
they are stored, so an invalid update leaves the existing object state unchanged.

```{r update_demography}
demography_updated <- demography
demography_updated[
  demography_updated[["region"]] == "north" & demography_updated[["age_group"]] == "18+",
  "population"
] <- 260000

region$set_demography(demography = demography_updated)

region %.% demography
```

```{r update_adjacency}
adjacency_updated <- adjacency
adjacency_updated[
  adjacency_updated[["from"]] == "north" & adjacency_updated[["to"]] == "south",
  "adjacency"
] <- 0.75
adjacency_updated[
  adjacency_updated[["from"]] == "south" & adjacency_updated[["to"]] == "north",
  "adjacency"
] <- 0.75

region$set_regions(regions = c("north", "south"))
region$set_adjacency(adjacency = adjacency_updated)

region %.% adjacency
```

# Validation

The module checks that selected regions are available in the stored demography and adjacency. For example, trying to
select a region that is not present in the data will fail.

```{r invalid_region, eval = FALSE}
region$set_regions(regions = "west")
```

The adjacency data must also contain the same region identifiers in the `from` and `to` columns.

```{r invalid_adjacency, eval = FALSE}
bad_adjacency <- data.frame(
  from = c("north", "south"),
  to = c("south", "east"),
  adjacency = c(1, 1)
)

region$set_adjacency(adjacency = bad_adjacency)
```


```{r cleanup, include = FALSE}
rm(
  adjacency,
  adjacency_updated,
  all_regions,
  demography,
  demography_updated,
  region
)
gc()
```
