---
title: "DiseasyRegions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DiseasyRegions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(diseasy)
```

```{r hidden_options, include = FALSE}
withr::local_options("diseasystore.verbose" = FALSE)
withr::local_options("tibble.print_min" = 5)
```

# Overview

The `DiseasyRegions` module defines the regional scope for the models.
In addition, it provides a common place to define the demography available for those regions,
and the adjacency between regions.

The module is intentionally generic. Region identifiers are treated as ordinary character values and are matched
exactly. This means that, e.g., the region `"north"` is not interpreted as a parent of `"north_subregion"`.

# Data structure

`DiseasyRegions` uses three inputs:

- `regions`: a `character` vector defining the selected region scope.
- `demography`: a `data.frame` with one region column, optional stratification columns, and one population column.
- `adjacency`: a `data.frame` describing connectedness between pairs of regions
  (see the [Regional adjacency](#regional-adjacency) section for further details).

The demography data must include the columns `region` and `population`. Any additional columns are treated as
stratification variables.

```{r demography_data}
demography_nordic
```

The adjacency data must include exactly the columns `from`, `to`, and `adjacency`.
The `from` and `to` columns should contain the same set of region identifiers.

```{r adjacency_data}
adjacency_meta_nordic
```

# Regional adjacency

`DiseasyRegions` builds on the concept of a degree of "connectedness" between the different regions.
Here, this "connectedness" is referred to as "adjacency" in alignment with other models.

Different measures for adjacency can be extracted from real-world data which have very different interpretations.
Adjacency is often described from information about "movement" such as mobility data or surveys
on how people move between regions. Alternatively adjacency can be derived from "infection" data such as genetic
phylogeny.

While these seem very similar, there are substantial differences in how these adjacencies are expressed numerically,
and how they should be implemented in disease spread models.
Our framework assumes a reality where disease is transported between regions by people returning to their own region
(travel or commuting), that is: there is no immigration or emigration.
If the supplied `adjacency` describes a movement matrix it is furthermore assumed that people travelling _to_ a place
transmits disease equally with people travelling _from_ a place.
Which in turns means that the resulting infection matrix used in the models will be symmetric.

In `diseasy`, the adjacency of `DiseasyRegions` are implemented in SEIR-like models and it helps to be clear on
how this adjacency is used to illustrate these differences.

To begin, let's define the "movement matrix" $\Phi$ between $R$ connected regions:

$$
\overline{\overline{\Phi}} = \begin{bmatrix}
\phi_{1,1} & \cdots & \phi_{1,R} \\
\vdots & \ddots & \vdots \\
\phi_{R,1} & \cdots & \phi_{R,R}
\end{bmatrix}
$$

If we interpret these adjacencies as "the average fraction of contacts between regions", we can diagram the SEIR compartments
of a two-region model as:

```{r SEIR-structure, echo = FALSE, fig.align = "center", fig.cap = "SEIR model overview for a two region model", out.width = "120%"}
knitr::include_graphics(
  devtools::package_file(
    "vignettes/articles/article_data/diseasy-regions-model-structure.png"
  )
)
```

With the following equations for the infectious compartments in region $1$
$$
\dot{I}_1 = \beta\left(
\underbrace{S_1 \phi_{1,1}}_\textrm{stay}  \underbrace{\phi_{1,1} I_1}_\textrm{stay} +
\underbrace{S_1 \phi_{1,1}}_\textrm{stay}  \underbrace{\phi_{1,2} I_2}_\textrm{travel} +
\underbrace{S_1 \phi_{2,1}}_\textrm{travel}\underbrace{\phi_{2,2} I_2}_\textrm{stay} +
\underbrace{S_1 \phi_{2,1}}_\textrm{travel}\underbrace{\phi_{2,1} I_1}_\textrm{travel}\right) -
r_I I_1
$$
where we divide the types of contacts into whether individuals stay in their own region or travel[^1] to other regions.

[^1]: Note that they do not immigrate to other regions when they travel as this would change their region index.

In general, we can express the dynamics in matrix form assuming $R$ different regions
$$
\dot{\overline{I}} = \beta\overline{S}\circ\overline{\overline{\Theta}}\cdot\overline{I} - r_I \overline{I}
$$

Where $\circ$ is the Hadamard product (element-wise multiplication).

The matrix elements of $\overline{\overline{\Theta}}$ contains the probability that individual interact across regions,
with elements $\Theta_{x,y} = \sum_z \phi_{z,x} \phi_{z,y}$.

These elements compute the probability that individuals in region $x$ travel to region $z$ AND individuals in region $y$
travel to region $z$.

For the 2-region system, the mixing can be expressed in matrix form as:
$$
\overline{\overline{\Theta}} = \begin{bmatrix}
  \phi_{1,1}^2 + \phi_{2,1}^2 & \phi_{1,1}\phi_{1,2} + \phi_{2,1}\phi_{2,2} \\
  \phi_{2,2}\phi_{2,1} + \phi_{1,2}\phi_{1,1} & \phi_{2,2}^2 + \phi_{1,2}^2
\end{bmatrix}
$$

In this formalism, the $\Theta$ which satisfies:
$$
\dot{I} = \dot{I}_1 + \dot{I}_2 = \beta S I - r_I I
$$

requires
$$
\overline{S}\cdot\overline{\overline{\Theta}}\cdot\overline{I} = S I
$$

That leads to the following expression:
$$
S_1 (\phi_{1,1}^2 + \phi_{2,1}^2) I_1 + S_1 (\phi_{1,1}\phi_{1,2} + \phi_{2,1}\phi_{2,2}) I_2 + \\
S_2 (\phi_{2,2}^2 + \phi_{1,2}^2) I_2 + S_2 (\phi_{2,2}\phi_{2,1} + \phi_{1,2}\phi_{1,1}) I_1
$$

If we assume there is no spatial structure and therefore all matrix elements are identical:
$\phi_{x,y} = \phi$

Then we can reduce the expression
$$
\Rightarrow
S_1 (2\phi^2) I_1 + S_1 (2\phi^2) I_2 +
S_2 (2\phi^2) I_2 + S_2 (2\phi^2) I_1
$$

$$
\Rightarrow
S_1 (2\phi^2) (I_1 + I_2) +
S_2 (2\phi^2) (I_1 + I_2)
$$

$$
\Rightarrow
(2\phi^2) (S_1 + S_2) (I_1 + I_2)
$$

$$
\Rightarrow
(2\phi^2) S I
$$

Which means that the neutral matrix elements for the 2-region system are:
$$
\phi = \frac{1}{\sqrt{2}}
$$

and in general for the n-region system[^2]:
$$
\phi = \frac{1}{\sqrt{n}}
$$

[^2]: Exercise left for the reader.

This "movement matrix" $\overline{\overline{\Phi}}$ can be estimated from information
on how individuals travel, such as from survey or mobility data.

Alternatively, we can consider the $\overline{\overline{\Theta}}$ matrix which we can
think of as a "infection-flow" matrix. This matrix identifies the flow of infections from
individuals residing in regions $x$ to individuals residing in regions $y$.
Such matrix may be estimated from more direct measurements of disease flow such as from
genetic data or contact tracing.

The $\Phi$ matrix will be normalised such that its row sums are 1.

`DiseasyRegions` can take either type of adjacency matrix.

The stored adjacency input is long form, but the module also exposes an `?DiseasyRegions$infection_flow_matrix` field.
This field returns a matrix representation of adjacency for the currently selected regions.

The matrix is derived in three steps:

- the stored adjacency data are filtered to the selected regions;
- the matrix is normalised so that row sums are 1.

# Creating a regional module

A `DiseasyRegions` object can be created and then configured with the selected regions, demography, and adjacency
data. Each setter validates the candidate input before storing it.

```{r create_module}
region <- DiseasyRegions$new()

region
```

```{r module_setters}
region$set_regions(regions = c("DK", "SE", "NO", "FI"))
region$set_demography(demography = demography_nordic)
region$set_adjacency(adjacency = adjacency_meta_nordic)

region
```

The `?DiseasyRegions$demography` field returns demography filtered to the currently selected regions.

```{r filtered_demography}
region %.% demography
```

Population totals can be computed directly from the filtered demography.

```{r population_totals}
aggregate(
  population ~ region,
  data = region %.% demography,
  FUN = sum
)
```

The active bindings `?DiseasyRegions$adjacency` and `?DiseasyRegions$infection_flow_matrix`
expose derived views of `adjacency` input after filtering to the selected regions.

```{r filtered_adjacency}
region %.% adjacency
```

```{r filtered_theta_matrix}
region %.% infection_flow_matrix
```

The module configuration can also be visualised via the `$plot()` method with or `plot()`.
```{r plot}
plot(region)
```

## Regional module with hierarchical structure (NUTS)

If you have demography and adjacency data for structured regions (such as NUTS), then `DiseasyRegionsNuts` can handle
the nested data.
```{r create_module_nuts}
region_nuts <- DiseasyRegionsNuts$new()

region_nuts
```

```{r nuts_module_setters}
region_nuts$set_regions(regions = "DK")
region_nuts$set_demography(demography = demography_nordic_nuts3)
region_nuts$set_adjacency(adjacency = adjacency_meta_nordic_nuts)

region_nuts
```

And again the module configuration can also be visualised with `plot()`.
```{r plot_nuts}
plot(region_nuts)
```

# Changing the selected regions

The selected regional scope can be changed with `$set_regions()`. The new regions must be available in the stored
adjacency and demography data.

```{r change_regions}
region$set_regions(regions = c("DK", "SE"))

region %.% regions
```

The active fields now reflect the updated scope.

```{r changed_demography}
region %.% demography
```

```{r changed_adjacency}
region %.% adjacency
```

```{r changed_theta_matrix}
region %.% infection_flow_matrix
```


# Validation

The module checks that selected regions are available in the stored demography and adjacency. For example, trying to
select a region that is not present in the data will fail.

```{r invalid_region}
try(region$set_regions(regions = "DE"))
```

The adjacency data must also contain the same region identifiers in the `from` and `to` columns.

```{r invalid_adjacency}
bad_adjacency <- data.frame(
  from = c("DE", "BE"),
  to = c("DE", "BE"),
  adjacency = c(1, 1)
)

try(region$set_adjacency(adjacency = bad_adjacency))
```


```{r cleanup, include = FALSE}
rm(
  adjacency,
  adjacency_new,
  all_regions,
  demography,
  demography_new,
  region
)
gc()
```
