---
title: "DiseasyVariant"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DiseasyVariant}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(diseasy)
```

```{r options, include = FALSE}
withr::local_options("digits" = 2, "tibble.print_min" = 5)
```

# Introduction
The `DiseasyVariant` is the module responsible for the implementation of variants.

# Configuring the module
The module can be initialized without setting any parameters.
```{r initialization}
variant <- DiseasyVariant$new()
```

By default, the module initializes with no variant scenario, and produces non-informative outputs.
```{r null_scenario}
variant$variants
```

To define a variant scenario, we manually add the variants of interest and their characteristics.
In the example below, we add the wild-type variant and a mutant variant named "Alpha".
```{r simple_configuration}
variant$add_variant(name = "WT")
variant$add_variant(
  name = "Alpha",
  characteristics = list(
    "relative_infection_risk" = 1.5,
    "introduction_date" = as.Date("2020-12-01")
  )
)

variant$variants
```

In the above example, there is perfect cross immunity between the variants (since we did not specify otherwise).
These interactions are in `DiseasyVariant` captured via `$cross_immunity`, a matrix showing the overlap in immunity
between the variants.

We can inspect the cross immunity matrix via `$cross_immunity`.
```{r simple_configuration_cross_immunity}
variant$cross_immunity
```

## Cross immunity interactions
If the variants have different levels of cross immunity, we can specify this in the `cross_immunity` characteristic
during initialisation.

Here we add a variant which has 85% cross immunity with the wild-type and the Alpha variant.
That is, when the Delta variant infects a host previously infected with the wild-type or Alpha variant, protection
provided the natural immunity in the host is reduced by 15%.

This cross immunity is asymmetric, meaning that when the wild-type or Alpha variant infects a host previously infected
with the Delta variant, the protection provided by the immunisation with the Delta variant is not reduced.
```{r cross_immunity_configuration}
variant$add_variant(
  name = "Delta",
  characteristics = list(
    "introduction_date" = as.Date("2021-05-01"),
    "cross_immunity" = list("WT" = 0.85, "Alpha" = 0.85)
  )
)

variant$cross_immunity
```


# Summarising the scenario
As with other `diseasy` modules, the `print()` generic can be used to get a human readable summary of the scenario.
```{r inspect_describe}
print(variant)
```
