---
title: "DiseasyDemography"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DiseasyDemography}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(diseasy)
```

```{r options, include = FALSE}
withr::local_options("digits" = 2, "tibble.print_min" = 5)
```

# Introduction
The `DiseasyDemography` is the module responsible for configuring the demography available to the model.

# Configuring the module
The module can be initialized without setting any parameters which configures the use of the default demography.
```{r initialization}
demography <- DiseasyDemography$new()

demography
```

This demography is pulled from the US Census Bureau and covers most of the world for 2020 but is not stratified beyond
the country-level.

If you have more specific demography information, you can change the configuration:
```{r user configuration}
demography$set_demography(demography = demography)
# which in this example does not actually change the configuration since
# `diseasy_demography` is used by default.

demography
```
