---
title: "DiseasyPopulation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DiseasyPopulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(diseasy)
```

```{r hidden_options, include = FALSE}
withr::local_options("diseasystore.verbose" = FALSE)
withr::local_options("tibble.print_min" = 5)
```

# Overview

The `DiseasyPopulation` module is responsible for defining the model population and interfacing with the models.

Presently, `DiseasyPopulation` combines `DiseasyRegions` and `DiseasyActivity` to configure a model population
stratified by age and geographical region. Where `DiseasyRegions` defines the _scope_ and possible stratifications of
the model population, `DiseasyPopulation` manages the stratification itself. Similarly, `DiseasyActivity` defines the
age-specific interactions between age groups as well as age-specific societal restrictions over time,
while `DiseasyPopulation` aggregates these effects into user stratified age groups.
Combined, this forms a model population stratified both spatially and by age groups.

# Defining the population of interest

The first step of defining the model population is to define a population of interest.

For `DiseasyPopulation`, this is done via `DiseasyRegions`[^1] which handles demography data
(see `vignette(diseasy-regions)` for more details).

[^1] Or, as in this case, `DiseasyRegionsNuts` which is a generalisation of `DiseasyRegions`.

```{r regions}
regions <- DiseasyRegionsNuts$new(
  demography = demography_nordic_nuts3, # Use demography data from EUROSTAT
  adjacency = adjacency_meta_nordic_nuts, # Use adjacency data from Meta
  regions = "DK" # Restrict population of interest to Danish population
)

regions
```

Once the scope of the population has been defined in `DiseasyRegions`, this can be passed to
`DiseasyPopulation` to handle the remaining configuration of the model population.

```{r load-regions}
population <- DiseasyPopulation$new(region = regions)

population
```

# Defining the age-specific interactions

If the model population should be stratified by age, the next step is to
configure the age-specific interactions between age groups (contact matrices).
In `diseasy`, this is done via the `DiseasyActivity`[^2].

[^2]: `DiseasyActivity` also defines (age-specific) changes in societal restrictions
if needed for the model -- but this is not done for this example.

```{r activity}
activity <- DiseasyActivity$new(
  contact_basis = contact_basis_nordic %.% DK # Contact matrices from `contactdata` package
)

activity
```

... and, once a basis of contact matrices has defined, the `DiseasyActivity`
can be passed to `DiseasyPopulation`.

```{r load-activity}
population$load_module(activity)

population
```

# Stratifying to form the model population

With the population basis defined via `DiseasyRegions` and `DiseasyActivity`,
the model population can now be stratified at various resolutions to before
being passed to the models.

To stratify by age, we use the `?DiseasyPopulation$stratify_age()` method:
```{r stratify-age}
population$stratify_age(age_cuts_lower = c(0, 20, 40, 60, 80))

# and we can inspect the defined demographic groups
population$group
```

To stratify by region, we use the `?DiseasyPopulation$stratify_regions()` method:
```{r stratify-regions}
population$stratify_regions(regional_stratification = "NUTS 2")

# .. and we can inspect the updated demographic groups
population$group
```

# Populations outside the Nordics

To keep the installation size of `diseasy` light, we only bundle population data
for the Nordic countries.

To generate data for other regions, see the data generating functions:
`generate_demography()`, `generate_contact_basis()`, etc.

```{r cleanup, include = FALSE}
rm(
  regions,
  activity,
  population
)
gc()
```
