---
title: "DiseasyModelOde"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DiseasyModelOde}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(diseasy)
```

```{r options, include = FALSE}
withr::local_options("diseasy.logging" = FALSE)
```

# Introduction
One of the model families in `diseasy` is the family of Ordinary Differential Equation (ODE) models.
This family of models is presently represented only by the `?DiseasyModelOdeSeir`, a model template that implements
SEIR (Susceptible, Exposed, Infected, Recovered) ODE models according to user specification.
This template allows for any number of $E$, $I$ or $R$ states in the model and interfaces with the `?DiseasySeason`,
`?DiseasyActivity`, `?DiseasyVariant` and `?DiseasyImmunity` modules for scenario definitions.

The workflow for creating a specific `?DiseasyModelOdeSeir` model would roughly follow these steps:

1) **Defining the scenario.**
  The first step to using most models in `diseasy` is to define the scenario. That is, to configure the functional
  modules that the model templates takes as argument (e.g. `?DiseasyActivity`).
  These steps provide the model template with information about the population, the activity, variants to include etc.
2) **Choosing the model configuration.**
  The model template may have some configurability that you control. In the case of `?DiseasyModelOdeSeir` for example,
  you must choose the number of consecutive $E$, $I$ and $R$ states in the model.
3) **Mapping to observables.**
  The ODE family of models is in some ways bespoke in the sense that the user needs provide a mapping from the
  observables to the "true" incidence of infections. The model is then initialised from this estimated incidence.
  If the model should provide predictions for observables other than the "true" incidence, the user needs to provide
  mapping between the model incidence and the observables in question.

These steps are outlined in individual sections below where we work through defining a model in a specific example.
Note that this example is contrived in the sense that the data we use in this example have been generated with this
exact model template and we therefore can match the observations near-perfectly.

# Example data
The data we will use this example is provided by `?DiseasystoreSeirExample` which represents data in a Danish context.

We first create an observables module to interface with this data:

```{r observables, eval = rlang::is_installed("RSQLite")}
# Configure a observables module with the example data and database
observables <- DiseasyObservables$new(
  diseasystore = DiseasystoreSeirExample,
  conn = DBI::dbConnect(RSQLite::SQLite())
)
```

The data is stratified by age and covers a single infection wave.
It contains three observables of interest for our example:

- *n_infected*: The true number of infected in the data.
- *n_positive*: The observed subset of infected
  (65% of true infected are observed but fewer are observed during the weekend).
- *n_admissions*: The number of new hospital admissions due to infections
  (risk of hospitalisation dependent on age group).

When running the models, we should also define at what point in time they should provide predictions for.
This is done by setting the `last_queryable_date` in the observables module:
```{r last_queryable_date, eval = rlang::is_installed("RSQLite")}
last_queryable_date <- observables$ds$min_start_date + 42
```

Lets first look at the data we have available:
```{r example data infected, eval = rlang::is_installed(c("ggplot2", "RSQLite")), fig.alt = "Data to model: True number of infected."}
observables$get_observation(
  observable = "n_infected",
  stratification = rlang::quos(age_group),
  start_date = observables$ds$min_start_date,
  end_date = observables$ds$max_end_date
) |>
  ggplot2::ggplot(ggplot2::aes(x = date, y = n_infected)) +
  ggplot2::geom_vline(
    xintercept = last_queryable_date,
    color = "black",
    linetype = "dashed"
  ) +
  ggplot2::geom_point(color = "deepskyblue3") +
  ggplot2::labs(title = "True number of infected") +
  ggplot2::facet_grid(~ age_group)
```

```{r example data positive, eval = rlang::is_installed(c("ggplot2", "RSQLite")), fig.alt = "Data to model: Number of positive cases."}
observables$get_observation(
  observable = "n_positive",
  stratification = rlang::quos(age_group),
  start_date = observables$ds$min_start_date,
  end_date = observables$ds$max_end_date
) |>
  ggplot2::ggplot(ggplot2::aes(x = date, y = n_positive)) +
  ggplot2::geom_vline(
    xintercept = last_queryable_date,
    color = "black",
    linetype = "dashed"
  ) +
  ggplot2::geom_point(color = "seagreen") +
  ggplot2::labs(title = "Observed number of positive cases") +
  ggplot2::facet_grid(~ age_group)
```

```{r example data admission, eval = rlang::is_installed(c("ggplot2", "RSQLite")), fig.alt = "Data to model: Number of hospital admissions."}
observables$get_observation(
  observable = "n_admission",
  stratification = rlang::quos(age_group),
  start_date = observables$ds$min_start_date,
  end_date = observables$ds$max_end_date
) |>
  ggplot2::ggplot(ggplot2::aes(x = date, y = n_admission)) +
  ggplot2::geom_vline(
    xintercept = last_queryable_date,
    color = "black",
    linetype = "dashed"
  ) +
  ggplot2::geom_point(color = "violetred3") +
  ggplot2::labs(title = "Number of admissions") +
  ggplot2::facet_grid(~ age_group)
```

# Defining the scenario
The first step to creating a `?DiseasyModelOdeSeir` model is to define the scenario.
In this case, as we stated above, we can exactly match the scenario to the data since
the data was created using this class of models.

First we define the activity scenario via the `?DiseasyActivity` module.
This scenario describes the changes in activity in the society (e.g. restrictions).
To see more on how to configure this module, see `vignette("diseasy-activity")`.

```{r diseasy_activity}
# Configure an activity module using Danish population and contact information.
activity <- DiseasyActivity$new()
activity$set_contact_basis(contact_basis = contact_basis$DK)
activity$set_activity_units(dk_activity_units)

# The level of activity is fixed to the "baseline" level throughout the simulation.
activity$change_activity(date = as.Date("2020-01-01"), opening = "baseline")
```

This particular scenario also include simple seasonality and waning immunity:
```{r diseasy_season}
# Waning immunity scenario
immunity <- DiseasyImmunity$new()
immunity$set_exponential_waning(time_scale = 180)

# Season scenario
season <- DiseasySeason$new()
season$set_reference_date(as.Date("2020-01-01"))
season$use_cosine_season()
```

# Choosing the model configuration
The `?DiseasyModelOdeSeir` model template has a number of hyperparameters and
parameters that should be configured for the model to run.

The hyperparameters are:

- *compartment_structure*: A named vector specifying the number of consecutive
  $E$, $I$ and $R$ states in the model.
- *age_cuts_lower*: A vector of age cuts for the age groups in the model.

The parameters are:

- *disease_progression_rates*: A named vector specifying the progression rates
  from $E$ to $I$ and from $I$ to $R$ (i.e. waiting times).
- *overall_infection_risk*: The overall risk of infection in the population
  (the overall scaling of infectivity).

```{r model configuration, eval = rlang::is_installed("RSQLite")}
# Hyper parameters
compartment_structure <- c("E" = 2L, "I" = 1L, "R" = 2L)
age_cuts_lower <- c(0, 30, 60)

# Parameters
disease_progression_rates <- c("E" = 1 / 2.1, "I" = 1 / 4.5)
overall_infection_risk <- 0.025

# When to predict?
observables$set_last_queryable_date(last_queryable_date)
```


# Mapping observables to model and back
Being a SEIR model, the `?DiseasyModelOdeSeir` model template runs the dynamics of the
infections in the population. These dynamics are inferred from the incidence of the
disease in the population.

Mathematically, we map from the set of real-world observables $\vec{O}(t)$ to the
incidence $I(t)$ via the map $f$:
$$\vec{O}(t) \overset{f}{\rightarrow} I(t)$$
The model then produces a number of observations under the given configuration.

By default, the model will predict the evolution of the incidence[^1], $\tilde{I}(t)$,
but other outputs can be configured (see further below).

[^1]: Measured as the rate of individuals exiting the first `I` state.

To get predictions for other real-world observables, we must map from the incidence
(or other output) to the observable of interest.

From the model incidence $\tilde{I}(t)$, we then map to the observable estimates
$\tilde{O}_i(t)$ via the maps $g_i$:
$$\tilde{I}(t) \overset{g_1}{\rightarrow} \tilde{O}_1(t)$$
$$\tilde{I}(t) \overset{g_2}{\rightarrow} \tilde{O}_2(t)$$
$$\vdots$$
$$\tilde{I}(t) \overset{g_n}{\rightarrow} \tilde{O}_n(t)$$

## Mapping from observables to incidence
For the model to be initialised, it needs an estimate of the incidence in the population.
This incidence is then used to infer the dynamics of the disease in the population
(See the article `SEIR-initialisation` in the online documentation for more information).

If an estimate of the incidence is not defined in the `diseasystore`, or if you wish to provide
your own estimate from the observable, we can define a "synthetic observable" in the
`DiseasyObservables` module for the model to use.

In this example, the incidence is not defined in the `diseasystore` and we therefore
define a synthetic observable for the incidence which maps from the number of positive
cases to the incidence.
In many cases, testing data may not be reliable, and you may want to infer the
incidence from other sources such as waste water data or admissions data.

```{r incidence map, eval = rlang::is_installed("RSQLite")}
# Map observables to incidence
observables$define_synthetic_observable(
  name = "incidence",
  mapping = \(n_positive, n_population) n_positive / (n_population * 0.65)
)
```


## Mapping from model incidence to observables
By default, the model can only return predictions for the true number of infected
(`n_infected`) and the incidence (`incidence`) since these are the only observables
"known" to the model.

Any additional observables must be mapped from the incidence provided by the model, or
other output must be defined to map from (see further below).

To convert from model outputs to real-world observables, we define a "map" and a
"reduce" function:

The *map* function is passed to a `dplyr::group_map()` call on a `data.frame` grouped
by the stratification.
The function should be a two-argument function where the first argument contains a
`data.frame` with columns `n_infected` and `population` (i.e. the size of the group),
and where the second argument contains the group (as a `data.frame`).
Note that the output of this function should contain the groups provided in the second
argument.

The *reduce* function is passed to a `dplyr::summarise()` call on the output of the
map. By default, a simple sum is used which works for all counting data.
If the observable is not a count, the user must provide a custom reduce function.
For example, the incidence is a rate and to summarise a weighted sum must be made.

In this example, we define maps from the incidence to the number of positive cases and
the number of hospital admissions. These are both counting observables and the default
*reduce* function can be used.

```{r observable maps}
# Maps between the internal model incidence and observables
map_to_n_positive <- list(
  "map" = \(.x, .y) {
    dplyr::transmute(
      .x,
      .data$date,
      "n_positive" = 0.65 * .x$n_infected
    )
  }
)

map_to_n_admission <- list(
  "map" = \(.x, .y) {
    # Risk per age group
    risk_of_admission <- c("00-29" = 0.001, "30-59" = 0.01, "60+" = 0.1)

    # Delay admissions by a fixed distribution
    delay_distribution <- c(0, 0, 0.2, 0.3, 0.3, 0.1, 0.1) # Must sum = 1

    data.frame(
      "date" = as.vector(
        outer(.x$date, seq_along(delay_distribution) - 1, "+")
      ),
      "n_admission" = as.vector(
        outer(
          .x$n_infected * risk_of_admission[.y$age_group],
          delay_distribution,
          "*"
        )
      )
    )
  }
)

# These maps are provided via a list to the model with the names of
# the elements denoting the observable being mapped to.
model_output_to_observable <- list(
  "n_positive" = map_to_n_positive,
  "n_admission" = map_to_n_admission
)
```

### Custom model output
The `diseasy` ODE model templates uses a generic system for generating observables
from the state_vector, $\overline{\psi}$.

A observable can be configured to either derive from either the full state vector
$\overline{\psi}$ or the infection matrix $f$ (see the article `SEIR-implementation`
in the online documentation).

In either case, the observable is configured by defining a vector of weights.
For each of such vectors, a "surveillance" state is added to the state_vector which
integrates the inner-product $\overline{\omega}_o \cdot \overline{\psi}$ or
$\overline{\omega}_o \cdot \sum_{rows}\overline{\overline{f}}$ depending on which
element the observable is set to track.

In the former case, the surveillance states derived from the state_vector where as
in the latter the observable is derived from the new infections occurring in the model.

In the case of the "n_infected" observable defined by default in the model, the weights
are essentially a matrix with weights only for each $I_1$ state[^2] and the observable
tracks the state_vector.

[^2]: The "n_infected" observable does not actually create the additional surveillance states in the implementation
since it can be computed directly from $\overline{\psi}$, but it used here as a illustrative example.


For example, in SEIR case ($N = L = M = 1$) with two age-groups, the weights take the
following form:

$$ \overline{\overline{\Omega}}_{n_{infected}} = \underbrace{\begin{bmatrix}
  0 &  r_i & 0 & 0 & 0 & 0 & 0 & 0 \\
  0 & 0 & 0 & 0 &  r_i & 0 & 0 & 0
\end{bmatrix}}_{\begin{matrix}
  E_1 & I_1 & R_1 & E_2 & I_2 & R_2 & S_1 & S_2
\end{matrix}}
$$
where $r_i$is the rate of transition out of $I$ (i.e. the reciprocal of the time spent
in $I$ state).

As stated above, the model always provides a prediction for the "n_infected" observable
but there are cases where other observables are automatically configured by `diseasy`.

For example, when the model includes a `?DiseasyImmunity` module, observables are
configured to track outcomes related to immunity.

Here, waning target named "hospitalisation" or "death" in the immunity scenario are
assigned their own observables which is derived from the infection matrix
(Note that there is no difference in waning between variants included in the model).

In the SEIR3 case ($N = L = 1, M = 3$) with two age-groups, the weights take the following form:
$$ \overline{\overline{\Omega}}_{immunity} = \underbrace{\begin{bmatrix}
  \gamma_1 & \gamma_2 & \gamma_3 & 0        & 0        & 0        & 1 & 0 \\
  0        & 0       & 0         & \gamma_1 & \gamma_2 & \gamma_3 & 0 & 1
\end{bmatrix}}_{\begin{matrix}
  \;\;R^1_1\; & R^2_1\; & R^3_1\;\; & R^1_2\; & R^2_2\; & R^3_2\;\; & S_1\; & S_2
\end{matrix}}
$$

In the SEIR2 case ($N = L = 1, M = 2$) with two age-groups and two variants, the weights take the following form:
$$ \overline{\overline{\Omega}}_{immunity} = \underbrace{\begin{bmatrix}
  \gamma_1 & \gamma_2 & 0        & 0        & \gamma_1 & \gamma_2 & 0        & 0        & 1 & 0 \\
  0        & 0        & \gamma_1 & \gamma_2 & 0        & 0        & \gamma_1 & \gamma_2 & 0 & 1
\end{bmatrix}}_{\begin{matrix}
  R^1_{1,1} & R^2_{1,1} & R^1_{2,1} & R^2_{2,1} & R^1_{1,2} & R^2_{1,2} & R^1_{2,2} & R^2_{2,2} & S_1 & S_2
\end{matrix}}
$$


# Putting it all together
With all the configuration done above, we can now create our model instance.

```{r model creation, eval = rlang::is_installed("RSQLite")}
model <- DiseasyModelOdeSeir$new(
  # Scenario configuration
  activity = activity,
  immunity = immunity,
  season = season,
  observables = observables,

  parameters = list(
    # Parameter and hyper-parameter configuration
    "compartment_structure" = compartment_structure,
    "age_cuts_lower" = age_cuts_lower,
    "overall_infection_risk" = overall_infection_risk,
    "disease_progression_rates" = disease_progression_rates,

    # Mapping to observables
    "model_output_to_observable" = model_output_to_observable
  )
)
```

# Predicting the incidence
From this point, we can easily get predictions for the individual observables.
```{r, eval = rlang::is_installed("RSQLite")}
model$get_results(
  observable = "incidence",
  prediction_length = 5
)
```

... and we can plot the predictions along with observations for the data
```{r, fig.alt = "Model prediction for the number of incidence - stratified by age group.", eval = rlang::is_installed("RSQLite")}
plot(
  model,
  observable = "incidence",
  prediction_length = 60
)
```

```{r configuration test - incidence, include = FALSE, eval = rlang::is_installed("RSQLite")}
# This code block is used as a sanity check that the models work as expected
# When the model works, we should be able to closely match model with data
# - when not, we get informed by an error
x <- model$get_results(observable = "incidence", prediction_length = 60)
y <- model$get_data(
  observable = "incidence", prediction_length = 60, period = "plotting"
) |>
  dplyr::filter(.data$t > 0)
stopifnot(
  "Model prediction varies more than expected from example data" = {
    purrr::every(abs(x$incidence - y$incidence) / x$incidence, ~ . < 0.4)
  }
)
```

The results are easily stratified by the stratifications in the data
```{r, eval = rlang::is_installed("RSQLite")}
model$get_results(
  observable = "incidence",
  prediction_length = 3,
  stratification = rlang::quos(age_group)
)
```

... and we can see the plots for each strata.
```{r, fig.alt = "Model prediction for the number of incidence - stratified by age group.", eval = rlang::is_installed("RSQLite")}
plot(
  model,
  observable = "incidence",
  prediction_length = 60,
  stratification = rlang::quos(age_group)
)
```


This can be done for all the observables we have mapped in the model.
```{r, eval = rlang::is_installed("RSQLite")}
model$get_results(
  observable = "n_infected",
  prediction_length = 5
)
```

```{r, fig.alt = "Model prediction for the number of infected - stratified by age group.", eval = rlang::is_installed("RSQLite")}
plot(
  model,
  observable = "n_infected",
  prediction_length = 60
)
```

```{r configuration test - n_infected, include = FALSE, eval = rlang::is_installed("RSQLite")}
# This code block is used as a sanity check that the models work as expected
# When the model works, we should be able to closely match model with data
# - when not, we get informed by an error
x <- model$get_results(observable = "n_infected", prediction_length = 60)
y <- model$get_data(
  observable = "n_infected", prediction_length = 60, period = "plotting"
) |>
  dplyr::filter(.data$t > 0)
stopifnot(
  "Model prediction varies more than expected from example data" = {
    purrr::every(abs(x$n_infected - y$n_infected) / x$n_infected, ~ . < 0.4)
  }
)
```

```{r, eval = rlang::is_installed("RSQLite")}
model$get_results(
  observable = "n_infected",
  prediction_length = 3,
  stratification = rlang::quos(age_group)
)
```


```{r, fig.alt = "Model prediction for the number of infected - stratified by age group.", eval = rlang::is_installed("RSQLite")}
plot(
  model,
  observable = "n_infected",
  prediction_length = 60,
  stratification = rlang::quos(age_group)
)
```


```{r, eval = rlang::is_installed("RSQLite")}
model$get_results(
  observable = "n_positive",
  prediction_length = 5
)
```

```{r, fig.alt = "Model prediction for the number of positive.", eval = rlang::is_installed("RSQLite")}
plot(
  model,
  observable = "n_positive",
  prediction_length = 60
)
```

```{r configuration test - n_positive, include = FALSE, eval = rlang::is_installed("RSQLite")}
# This code block is used as a sanity check that the models work as expected
# When the model works, we should be able to closely match model with data
# - when not, we get informed by an error
x <- model$get_results(observable = "n_positive", prediction_length = 60)
y <- model$get_data(
  observable = "n_positive", prediction_length = 60, period = "plotting"
) |>
  dplyr::filter(.data$t > 0)
stopifnot(
  "Model prediction varies more than expected from example data" = {
    purrr::every(abs(x$n_positive - y$n_positive) / x$n_positive, ~ . < 0.4)
  }
)
```

```{r, eval = rlang::is_installed("RSQLite")}
model$get_results(
  observable = "n_positive",
  prediction_length = 3,
  stratification = rlang::quos(age_group)
)
```

```{r, fig.alt = "Model prediction for the number of positive - stratified by age group.", eval = rlang::is_installed("RSQLite")}
plot(
  model,
  observable = "n_positive",
  prediction_length = 60,
  stratification = rlang::quos(age_group)
)
```

```{r, eval = rlang::is_installed("RSQLite")}
model$get_results(
  observable = "n_admission",
  prediction_length = 5
)
```

```{r, fig.alt = "Model prediction for the number of hospital admissions.", eval = rlang::is_installed("RSQLite")}
plot(
  model,
  observable = "n_admission",
  prediction_length = 60
)
```

```{r configuration test - n_admission, include = FALSE, eval = rlang::is_installed("RSQLite")}
# This code block is used as a sanity check that the models work as expected
# When the model works, we should be able to closely match model with data
# - when not, we get informed by an error
x <- model$get_results(observable = "n_admission", prediction_length = 60)
y <- model$get_data(
  observable = "n_admission", prediction_length = 60, period = "plotting"
) |>
  dplyr::filter(.data$t > 0)
stopifnot(
  "Model prediction varies more than expected from example data" = {
    purrr::every(abs(x$n_admission - y$n_admission) / x$n_admission, ~ . < 0.4)
  }
)
```

```{r, eval = rlang::is_installed("RSQLite")}
model$get_results(
  observable = "n_admission",
  prediction_length = 3,
  stratification = rlang::quos(age_group)
)
```

```{r, fig.alt = "Model prediction for the number of hospital admissions - stratified by age group.", eval = rlang::is_installed("RSQLite")}
plot(
  model,
  observable = "n_admission",
  prediction_length = 60,
  stratification = rlang::quos(age_group)
)
```


```{r cleanup, include = FALSE}
if (exists("model")) rm(model)
gc()
```
