---
title: "DiseasyPopulation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DiseasyPopulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(diseasy)
```

```{r hidden_options, include = FALSE}
withr::local_options("diseasystore.verbose" = FALSE)
withr::local_options("tibble.print_min" = 5)
```

# Overview

The `DiseasyPopulation` module is responsible for defining the model population and interfacing with the models.

Presently, `DiseasyPopulation` combines `DiseasyRegions` and `DiseasyActivity` to configure a model population
stratified by age and geographical region. Where `DiseasyRegions` defines the _scope_ and possible stratifications of
the model population, `DiseasyPopulation` manages the stratification itself. Similarly, `DiseasyActivity` defines the
age-specific interactions between age groups as well as age-specific societal restrictions over time,
while `DiseasyPopulation` aggregates these effects into user stratified age groups.
Combined, this forms a model population stratified both spatially and by age groups.

# Defining the population of interest

The first step of defining the model population is to define a population of interest.

For `DiseasyPopulation`, this is done via `DiseasyRegions`[^1] which handles demography data
(see `vignette(diseasy-regions)` for more details).

[^1]: Or, as in this case, `DiseasyRegionsNuts` which is a generalisation of `DiseasyRegions`.

```{r regions}
regions <- DiseasyRegionsNuts$new(
  demography = demography_nordic_nuts3, # Use demography data from EUROSTAT
  adjacency = adjacency_meta_nordic_nuts3, # Use adjacency data from Meta
  area = "DK" # Restrict population of interest to Danish population
)

regions
```

Once the scope of the population has been defined in `DiseasyRegions`, this can be passed to
`DiseasyPopulation` to handle the remaining configuration of the model population.

```{r load-regions}
population <- DiseasyPopulation$new(regions = regions)

population
```

# Defining the age-specific interactions

If the model population should be stratified by age, the next step is to
configure the age-specific interactions between age groups (contact matrices).
In `diseasy`, this is done via the `DiseasyActivity`[^2].

[^2]: `DiseasyActivity` also defines (age-specific) changes in societal restrictions
if needed for the model -- but this is not done for this example.

```{r activity}
activity <- DiseasyActivity$new(
  contact_basis = contact_basis_nordic %.% DK
)

activity
```

... and, once a basis of contact matrices has defined, the `DiseasyActivity`
can be passed to `DiseasyPopulation`.

```{r load-activity}
population$load_module(activity)

population
```

# Stratifying to form the model population

With the population basis defined via `DiseasyRegions` and `DiseasyActivity`,
the model population can now be stratified at various resolutions to before
being passed to the models.

To stratify by age, we use the `?DiseasyPopulation$stratify_age()` method:
```{r stratify-age}
population$stratify_age(age_cuts_lower = c(0, 20, 40, 60, 80))

# and we can inspect the defined demographic groups
population$groups
```

To stratify by region, we use the `?DiseasyPopulation$stratify_regions()` method:
```{r stratify-regions}
population$stratify_regions(regional_stratification = "NUTS 2")

# .. and we can inspect the updated demographic groups
population$groups
```

# Populations outside the Nordics

To keep the installation size of `diseasy` light, we only bundle population data
for the Nordic countries.

To generate data for other regions, see the data generating functions:
`generate_demography()`, `generate_contact_basis()`, etc.



# Notes: Regional aggregation of contact matrices

The source data for the contact matrices we use in `diseasy` is typically collections of survey data where
participants have been instructed to count contacts with other individuals along with the approximate age of
these individuals. These survey data are then corrected for differences in demography of the participants
and the group they are meant to represent (e.g. a given country) for form a "full" contact matrix for the
target population.


```{r, include = FALSE}
# Define some helper functions

# A plotter for the adjacency graphs
adjacency_graph <- function(adjacency) {
  graph <- igraph::graph_from_adjacency_matrix(
    adjmatrix = adjacency,
    mode = "undirected",
    diag = FALSE
  )

  region <- stringr::str_extract(population, r"{\w+$}")
  age <- stringr::str_extract(population, r"{^\w+}")

  layout <- cbind(
    x = as.numeric(factor(region)),
    y = as.numeric(factor(age))
  )

  # Add some jitter so vertices in the same region/age group do not overlap
  layout <- layout + matrix(
    stats::runif(length(layout), -0.3, 0.3),
    ncol = 2
  )

  plot(
    graph,
    layout = layout,
    vertex.color = factor(region),
    vertex.shape = "circle",
    vertex.label = population,
    vertex.size = 30
  )
}

# A method for aggregating data (typically adjacency) in new sub groups
contact_matrix <- function(adjacency, population, groups) {

  p <- purrr::map(
    unique(groups),
    ~ as.numeric(groups == .x)
  ) |>
    purrr::reduce(
      rbind,
      .init = matrix(0, nrow = 0, ncol = length(population))
    )

  # Count population by group
  n_groups <- lengths(split(groups, factor(groups)))[unique(groups)]

  colnames(p) <- population
  rownames(p) <- paste0(names(n_groups), " (", n_groups, ")")

  return(p %*% adjacency %*% t(p))
}

# A printing function which uses dots (...) to prevent wide and long matrices in vignette
print_truncated <- function(x, n = 4) {
  nr <- nrow(x)
  nc <- ncol(x)

  rows <- unique(c(seq_len(min(n, nr)), seq.int(max(nr - n + 1, 1), nr)))
  cols <- unique(c(seq_len(min(n, nc)), seq.int(max(nc - n + 1, 1), nc)))

  y <- format(x[rows, cols, drop = FALSE])

  if (nr > 2 * n)
    y <- rbind(y[seq_len(n), , drop = FALSE], "...", y[(n + 1):(2 * n), , drop = FALSE])

  if (nc > 2 * n)
    y <- cbind(y[, seq_len(n), drop = FALSE], "...", y[, (n + 1):(2 * n), drop = FALSE])

  print(y, quote = FALSE)
}
```


To gain some intuition on the how these contact matrices can be interpreted and manipulated we will
begin by considering some toy-examples with a toy population of age groups ("A", "B", "C") and spatial
regions ("x", "y", "z", "w").

```{r}
population <- c(
  rep("A/x", 3), rep("B/x", 2), rep("C/x", 1),
  rep("A/y", 1), rep("B/y", 2), rep("C/y", 3),
  rep("A/z", 1), rep("B/z", 1), rep("C/z", 1),
  rep("A/w", 1), rep("B/w", 2), rep("C/w", 1)
)
regions <- unique(stringr::str_replace(population, "[A-Z]", "*"))
age_groups <- unique(stringr::str_replace(population, "[a-z]", "*"))

print(lengths(split(population, factor(population))[unique(population)]))
```

with age-specific per-capita contact rates ("C" domain) of:
```{r}
c_matrix <- sqrt(outer(1 / seq_along(age_groups), 1 / seq_along(age_groups)))
dimnames(c_matrix) <- list(age_groups, age_groups)
print(round(c_matrix, digits = 2))
```

and corresponding total contacts ("T" domain):
```{r}
N_age <- lengths(
  split(
    stringr::str_extract(population, "[A-Z]"),
    factor(stringr::str_extract(population, "[A-Z]"))
  )
)

t_matrix <- round(c_matrix * tcrossprod(N_age))
print(t_matrix)
```

## A well-mixed example

Consider the well-mixed case where a population of individuals across different age groups and
spatial locations are equally likely to be connected.

We can draw a realisation of such a network from our contact matrices
```{r}
# First we need a implementation of `sample` which is not moronic
# - Just sample from the vector and by "sample" I mean "sample" not "permutate"
sample <- \(x, size) x[sample.int(length(x), size, replace = TRUE)]

# Convert T-matrix to long-form data.frame
one_way_links <- tidyr::expand_grid(
  i_age_group = age_groups,
  j_age_group = age_groups
) |>
  dplyr::mutate(
    "n_contacts" = as.vector(t_matrix)
  ) |>
  tidyr::uncount(.data$n_contacts) |>
  dplyr::select(!"n_contacts")

# Randomly draw i indicies (from nodes)
one_way_links <- one_way_links |>
  dplyr::group_by(.data$i_age_group) |>
  dplyr::group_modify(
    ~ .x |>
      dplyr::mutate(
        "i" = sample(
          x = which(
            stringr::str_starts(population, .y$i_age_group)
          ),
          size = dplyr::n()
        )
      )
  )

# Randomly draw j indices (to nodes)
one_way_links <- one_way_links |>
  dplyr::group_by(.data$i, .data$j_age_group) |>
  dplyr::group_modify(
    ~ .x |>
      dplyr::mutate(
        "j" = sample(
          x = which(
            stringr::str_starts(population, .y$j_age_group) &
              seq_along(population) != .y$i
          ),
          size = dplyr::n()
        )
      )
  )

# Finalise
one_way_links <- one_way_links |>
  dplyr::mutate( # Make lower
    "i_u" = pmax(.data$i, .data$j),
    "j_u" = pmin(.data$i, .data$j)
  ) |>
  dplyr::group_by(.data$i_u, .data$j_u) |>
  dplyr::summarise("n" = dplyr::n(), .groups = "drop") # Count connections between nodes

# Convert to adjacency matrix with 2-way interactions
adjacency_sw <- as.matrix(
  Matrix::sparseMatrix(
    i = one_way_links$i_u,
    j = one_way_links$j_u,
    x = one_way_links$n,
    dims = c(length(population), length(population)),
    dimnames = list(population, population)
  )
)

adjacency_sw <- (adjacency_sw + t(adjacency_sw)) / 2

print_truncated(adjacency_sw)
```


```{r, eval = rlang::is_installed("igraph")}
adjacency_graph(adjacency_sw)
```

From this full adjacency graph, we can project the contacts ("T" domain) into contact matrices in different stratifications.

Number of contacts stratified by age group and region (with population size in parentheses):
```{r}
T_age_region <- contact_matrix(
  adjacency = adjacency_sw,
  population = population,
  groups = population # No aggregation
)

print_truncated(T_age_region)
cat("Total contacts: ", sum(T_age_region))
```

Stratify only by region:
```{r}
T_region <- contact_matrix(
  adjacency = adjacency_sw,
  population = population,
  groups = stringr::str_replace(population, r"{[A-Z]}", "*")
)

print(T_region)
cat("Total contacts: ", sum(T_region))
```

Stratify only by age group (and compare with input used to generate the adjacency graph)
```{r}
T_age <- contact_matrix(
  adjacency = adjacency_sw,
  population = population,
  groups = stringr::str_replace(population, r"{[a-z]}", "*")
)

print(T_age)
cat("Total contacts: ", sum(T_age))

# Reference
print(t_matrix)
```

Lets see if we can achieve these same contacts without generating the network.

First, the mixing matrix for well-mixed scenario
```{r}
mixing_sw <- matrix(
  data = 1,
  nrow = length(regions),
  ncol = length(regions),
  dimnames = list(regions, regions)
)
print(mixing_sw)
```

We compute the full per-capita contact matrix, $\overline{\overline{C}}$, from the age-specific contact matrix, $\overline{\overline{C}}_{age}$, the mixing matrix, $\overline{\overline{\Theta}}$, and the population vector, $\overline{N}$, via the relation:

$$
\frac{\overline{\overline{C}}}{\overline{N}\overline{\overline{1}}\overline{N}} = \frac{\overline{\overline{\Theta}} \otimes \overline{\overline{C}}_{age}}{\overline{N}(\overline{\overline{\Theta}} \otimes \overline{\overline{1}}_{age})\overline{N}}
$$
where $\otimes$ is the Kronecker product, and $\overline{\overline{1}}$ and $\overline{\overline{1}}_{age}$ are matrices of ones with same dimensions of $\overline{\overline{C}}$ and $\overline{\overline{C}}_{age}$ respectively.




$$
\overline{\overline{C}} = \left(\frac{\overline{N}\overline{\overline{1}}\overline{N}}{\overline{N}(\overline{\overline{\Theta}} \otimes \overline{\overline{1}}_{age})\overline{N}}\overline{\overline{\Theta}}\right)\otimes \overline{\overline{C}}_{age}
$$

Or, in simplified form:
$$
\overline{\overline{C}} = \overline{\overline{\Theta^\ast}} \otimes \overline{\overline{C}}_{age}
$$
Where
$$
\overline{\overline{\Theta^\ast}} = \frac{\overline{\overline{\Theta}}}{\overline{n}^T\overline{\overline{\Theta}}\overline{n}}
$$
with $\overline{n}$ being the population fraction:
$$
\overline{n} = \frac{\overline{N}}{\overline{1}^T\overline{N}}
$$
The normalizing constant is:
$$
\overline{n}^T\overline{\overline{\Theta}}\overline{n}
$$
This is the *population-pair-weighted* average of the entries of $\Theta$.

Consider a case where P and Q are the mixing-group labels of two independently selected people.
In this case,
$$
\overline{n}^T\overline{\overline{\Theta}}\overline{n} = \mathbb{E}[\Theta_{P,Q}].
$$
Therefore,
$$
\sum_{i,j} n_i n_j \Theta_{p,q}^\ast = 1.
$$
So the normalized mixing matrix has population-weighted mean equal to one.


The normalized entry $\Theta^{\ast}_{ij}$ is a relative contact multiplier:
For example,

$$
\Theta^\ast_{p,q} = 1.4
$$

means that contacts between mixing groups p and q are multiplied by 1.4, relative to the average mixing effect.


Let's see that computation in practice:
```{r}
N_regions <- matrix(
  lengths(
    split(
      stringr::str_replace(population, "[A-Z]", "*"),
      factor(stringr::str_replace(population, "[A-Z]", "*"))
    )[unique(stringr::str_replace(population, "[A-Z]", "*"))]
  ),
  ncol = 1,
  dimnames = list(regions, "n")
)
n_regions <- N_regions / sum(N_regions)

print(n_regions)
```

```{r}
tm <- mixing_sw
cm <- c_matrix

tm_norm <- tm / drop(t(n_regions) %*% tm %*% n_regions)

c_matrix_full_sw <- kronecker(tm_norm, cm)
dimnames(c_matrix_full_sw) <- list(unique(population), unique(population))
print_truncated(round(c_matrix_full_sw, digits = 1))
```

The per-capita contact matrix can then be cast to the number of contacts ("T" domain).
```{r}
N <- matrix(
  lengths(split(population, factor(population))[unique(population)]),
  ncol = 1
)

t_matrix_full_sw <- c_matrix_full_sw * tcrossprod(N)
print_truncated(round(t_matrix_full_sw, digits = 1))
```
Finally, we can aggregate the raw contact matrix in the same groups as the original reference (i.e. age groups only) and compare with that reference.
```{r}
T_age_indirect <- contact_matrix(
  t_matrix_full_sw,
  population = unique(population),
  groups = stringr::str_replace(unique(population), r"{[a-z]}", "*")
)

print(round(T_age_indirect, digits = 1))
cat("Total contacts: ", round(sum(T_age_indirect), digits = 1))

# Reference
print(t_matrix)
```

## A silo example
Consider a complementary network where a population of individuals across different age groups and
spatial locations are well-mixed only within their spatial region (a extremely "siloed" network).

We can draw a realisation of such a network from our contact matrices
```{r}
# Convert T matrix to long-form data.frame
one_way_links <- tidyr::expand_grid(
  i_age_group = age_groups,
  j_age_group = age_groups
) |>
  dplyr::mutate(
    "n_contacts" = as.vector(t_matrix)
  ) |>
  tidyr::uncount(.data$n_contacts) |>
  dplyr::select(!"n_contacts")

# Randomly draw i indicies (from nodes)
# verify they connect to j nodes in the region
# -- else re-draw from accepted samples
one_way_links <- one_way_links |>
  dplyr::group_by(.data$i_age_group) |>
  dplyr::group_modify(
    ~ {
      i_sampled <- .x |>
        dplyr::mutate(
          "i_age_group" = .y$i_age_group,
          "i" = sample(
            x = which(
              stringr::str_detect(population, .y$i_age_group)
            ),
            size = dplyr::n()
          ),
          "i_lab" = population[.data$i],
          "i_region" = stringr::str_extract(.data$i_lab, "[a-z]")
        )

      # Determine reachable connections
      i_sampled <- i_sampled |>
        dplyr::mutate(
          "any_j_reachable" = purrr::pmap_lgl(
            dplyr::select(i_sampled, "j_age_group", "i_region", "i"),
            \(j_age_group, i_region, i) {
              any(
                stringr::str_detect(population, stringr::str_replace(j_age_group, r"{\*}", i_region)) &
                  seq_along(population) != i
              )
            }
          )
        )

      # Determine reachable connections
      i_accepted <- i_sampled |>
        dplyr::filter(.data$any_j_reachable)

      # Re-sample unreachable from reachable
      i_resampled <- i_sampled |>
        dplyr::filter(!.data$any_j_reachable) |>
        dplyr::count(.data$i_age_group, .data$j_age_group) |>
        purrr::pmap(
          \(i_age_group, j_age_group, n) {
            i_accepted |>
              dplyr::filter(.data$i_age_group == {{ i_age_group }}, .data$j_age_group == {{ j_age_group }}) |>
              dplyr::slice_sample(n = {{ n }})
          }
        ) |>
        purrr::list_rbind()

      return(dplyr::select(rbind(i_accepted, i_resampled), !"i_age_group"))
    }
  )

# Randomly draw j indicies (to nodes) within the same region
one_way_links <- one_way_links |>
  dplyr::group_by(.data$i, .data$j_age_group, .data$i_region) |>
  dplyr::group_modify(
    ~ .x |>
      dplyr::mutate(
        "j" = sample(
          x = which(
            stringr::str_detect(population, stringr::str_replace(.y$j_age_group, r"{\*}", .y$i_region)) & seq_along(population) != .y$i
          ),
          size = dplyr::n()
        ),
        "j_lab" = population[.data$j],
        "j_region" = stringr::str_extract(.data$j_lab, "[a-z]")
      )
  )

# Finalise
one_way_links <- one_way_links |>
  dplyr::mutate( # Make lower
    "i_u" = pmax(.data$i, .data$j),
    "j_u" = pmin(.data$i, .data$j)
  ) |>
  dplyr::group_by(.data$i_u, .data$j_u) |>
  dplyr::summarise("n" = dplyr::n(), .groups = "drop") # Count connections between nodes

# Convert to adjacency matrix with 2-way interactions
adjacency_silo <- as.matrix(
  Matrix::sparseMatrix(
    i = one_way_links$i_u,
    j = one_way_links$j_u,
    x = one_way_links$n,
    dims = c(length(population), length(population)),
    dimnames = list(population, population)
  )
)

adjacency_silo <- (adjacency_silo + t(adjacency_silo)) / 2

print_truncated(adjacency_silo)
```

```{r, eval = rlang::is_installed("igraph")}
adjacency_graph(adjacency_silo)
```


Number of contacts stratified by age group and region
```{r}
T_age_region <- contact_matrix(
  adjacency = adjacency_silo,
  population = population,
  groups = population # No aggregation
)

print_truncated(T_age_region)
cat("Total contacts: ", sum(T_age_region))
```

Stratify only by region
```{r}
T_region <- contact_matrix(
  adjacency = adjacency_silo,
  population = population,
  groups = stringr::str_replace(population, r"{[A-Z]}", "*")
)

print(T_region)
cat("Total contacts: ", sum(T_region))
```

Stratify only by age group
```{r}
T_age <- contact_matrix(
  adjacency = adjacency_silo,
  population = population,
  groups = stringr::str_replace(population, r"{[a-z]}", "*")
)

print(T_age)
cat("Total contacts: ", sum(T_age))

# Reference
print(t_matrix)
```


Again, lets see if we can achieve these same contacts without generating the network.

```{r}
mixing_silo <- diag(
  x = 1,
  nrow = length(regions),
  ncol = length(regions)
)
dimnames(mixing_silo) <- list(regions, regions)
print(mixing_silo)
```


```{r}
tm <- mixing_silo
cm <- c_matrix

tm_norm <- tm / drop(t(n_regions) %*% tm %*% n_regions)

c_matrix_full_silo <- kronecker(tm_norm, cm)
dimnames(c_matrix_full_silo) <- list(unique(population), unique(population))
print_truncated(round(c_matrix_full_silo, digits = 1))
```

```{r}
t_matrix_full_silo <- c_matrix_full_silo * tcrossprod(N)
print_truncated(round(t_matrix_full_silo, digits = 1))
```

```{r}
T_age_indirect <- contact_matrix(
  t_matrix_full_silo,
  population = unique(population),
  groups = stringr::str_replace(unique(population), r"{[a-z]}", "*")
)

print(round(T_age_indirect, digits = 1))
cat("Total contacts: ", round(sum(T_age_indirect), digits = 1))

# Reference
print(t_matrix)
```

## A mixture example

Finally, lets test the indirect method on a mixture of the two examples:

```{r}
mixing_mix <- matrix(
  data = 0.2,
  nrow = length(regions),
  ncol = length(regions)
) +
  diag(
    x = 0.8,
    nrow = length(regions),
    ncol = length(regions)
  )

dimnames(mixing_mix) <- list(regions, regions)
print(mixing_mix)
```


```{r}
tm <- mixing_mix
cm <- c_matrix

tm_norm <- tm / drop(t(n_regions) %*% tm %*% n_regions)

c_matrix_full_mix <- kronecker(tm_norm, cm)
dimnames(c_matrix_full_mix) <- list(unique(population), unique(population))
print_truncated(round(c_matrix_full_mix, digits = 1))
```

```{r}
t_matrix_full_mix <- c_matrix_full_mix * tcrossprod(N)
print_truncated(round(t_matrix_full_mix, digits = 1))
```

```{r}
T_age_indirect <- contact_matrix(
  t_matrix_full_mix,
  population = unique(population),
  groups = stringr::str_replace(unique(population), r"{[a-z]}", "*")
)

print(round(T_age_indirect, digits = 1))
cat("Total contacts: ", round(sum(T_age_indirect), digits = 1))

# Reference
print(t_matrix)
```

```{r cleanup, include = FALSE}
rm(
  regions,
  activity,
  population
)
gc()
```
