---
title: "DiseasyPopulation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DiseasyPopulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(diseasy)
```

```{r hidden_options, include = FALSE}
withr::local_options("diseasystore.verbose" = FALSE)
withr::local_options("tibble.print_min" = 5)
```

# Overview

The `DiseasyPopulation` module is responsible for defining the model population and interfacing with the models.

Presently, `DiseasyPopulation` combines `DiseasyRegions` and `DiseasyActivity` to configure a model population
stratified by age and geographical region. Where `DiseasyRegions` defines the _scope_ and possible stratifications of
the model population, `DiseasyPopulation` manages the stratification itself. Similarly, `DiseasyActivity` defines the
age-specific interactions between age groups as well as age-specific societal restrictions over time,
while `DiseasyPopulation` aggregates these effects into user stratified age groups.
Combined, this forms a model population stratified both spatially and by age groups.

# Defining the population of interest

The first step of defining the model population is to define a population of interest.

For `DiseasyPopulation`, this is done via `DiseasyRegions`[^1] which handles demography data
(see `vignette(diseasy-regions)` for more details).

[^1] Or, as in this case, `DiseasyRegionsNuts` which is a generalisation of `DiseasyRegions`.

```{r regions}
regions <- DiseasyRegionsNuts$new(
  demography = demography_nordic_nuts3, # Use demography data from EUROSTAT
  adjacency = adjacency_meta_nordic_nuts3, # Use adjacency data from Meta
  area = "DK" # Restrict population of interest to Danish population
)

regions
```

Once the scope of the population has been defined in `DiseasyRegions`, this can be passed to
`DiseasyPopulation` to handle the remaining configuration of the model population.

```{r load-regions}
population <- DiseasyPopulation$new(regions = regions)

population
```

# Defining the age-specific interactions

If the model population should be stratified by age, the next step is to
configure the age-specific interactions between age groups (contact matrices).
In `diseasy`, this is done via the `DiseasyActivity`[^2].

[^2]: `DiseasyActivity` also defines (age-specific) changes in societal restrictions
if needed for the model -- but this is not done for this example.

```{r activity}
activity <- DiseasyActivity$new(
  contact_basis = contact_basis_nordic %.% DK
)

activity
```

... and, once a basis of contact matrices has defined, the `DiseasyActivity`
can be passed to `DiseasyPopulation`.

```{r load-activity}
population$load_module(activity)

population
```

# Stratifying to form the model population

With the population basis defined via `DiseasyRegions` and `DiseasyActivity`,
the model population can now be stratified at various resolutions to before
being passed to the models.

To stratify by age, we use the `?DiseasyPopulation$stratify_age()` method:
```{r stratify-age}
population$stratify_age(age_cuts_lower = c(0, 20, 40, 60, 80))

# and we can inspect the defined demographic groups
population$group
```

To stratify by region, we use the `?DiseasyPopulation$stratify_regions()` method:
```{r stratify-regions}
population$stratify_regions(regional_stratification = "NUTS 2")

# .. and we can inspect the updated demographic groups
population$groups
```

# Populations outside the Nordics

To keep the installation size of `diseasy` light, we only bundle population data
for the Nordic countries.

To generate data for other regions, see the data generating functions:
`generate_demography()`, `generate_contact_basis()`, etc.



# Notes: Regional aggregation of contact matrices
The source data for the contact matrices we use in `diseasy` is typically collections of survey data where
participants have been instructed to count contacts with other individuals along with the approximate age of
these individuals. These survey data are then corrected for differences in demography of the participants
and the group they are meant to represent (e.g. a given country) for form a "full" contact matrix for the
target population.

To gain some intuition on the how these contact matrices can be interpreted and manipulated we will
begin by considering some toy-examples with a toy population of age groups ("A", "B", "C") and spatial
regions ("x", "y").

```{r}
population <- c(
  rep("A/x", 1), #rep("B/x", 2), rep("C/x", 3),
  rep("A/y", 2), #rep("B/y", 2), rep("C/y", 1),
  rep("A/z", 1), #rep("B/z", 1), rep("C/z", 1),
  rep("A/w", 6)#, rep("B/w", 6), rep("C/w", 6)
)
regions <- unique(stringr::str_replace(population, "[A-Z]", "*"))
age_groups <- unique(stringr::str_replace(population, "[a-z]", "*"))

print(lengths(split(population, factor(population))[unique(population)]))
```

with per-capita contact rates ("C" domain) of:
```{r}
age_groups <- unique(stringr::str_replace(population, "[a-z]", "*"))
c_matrix <- sqrt(outer(1 / seq_along(age_groups), 1 / seq_along(age_groups)))
dimnames(c_matrix) <- list(age_groups, age_groups)
print(round(c_matrix, digits = 2))
```
and corresponding total contacts ("T" domain):
with per-capita contact rates ("C" domain) of:
```{r}
N <- lengths(
  split(
    stringr::str_extract(population, "[A-Z]"), 
    factor(stringr::str_extract(population, "[A-Z]"))
  )
)

N_effective <- tibble::tibble(
  "population" = population
) |>
  tidyr::separate_wider_delim(
    cols = "population", 
    delim = "/", 
    names = c("age_group", "region")
  ) |>
  dplyr::count(.data$age_group, .data$region) |>
  dplyr::summarise("avg_n" = mean(n), .by = "age_group") |>
  tibble::deframe()

t_matrix <- round(c_matrix * outer(N, N))
print(t_matrix)
```

## A well-mixed example
Consider a the fully-connected case where a population of individuals across different age groups and
spatial locations are equally likely to be connected.

We can draw a realisation of such a network from our contact matrices
```{r}
# First we need a implementation of `sample` which is not moronic
# - Just sample from the vector and by "sample" I mean "sample" not "permutate"
sample <- \(x, size) x[sample.int(length(x), size, replace = TRUE)]

one_way_links <- tidyr::expand_grid(
  i_age_group = age_groups,
  j_age_group = age_groups
) |>
  dplyr::mutate(
    "n_contacts" = as.vector(t_matrix)
  ) |>
  tidyr::uncount(.data$n_contacts) |>
  dplyr::select(!"n_contacts") |>
  dplyr::group_by(.data$i_age_group) |>
  dplyr::group_modify(
    ~ .x |>
      dplyr::mutate(
        "i" = sample(
          x = which(
            stringr::str_starts(population, .y$i_age_group)
          ), 
          size = dplyr::n()
        )
      )
  ) |> 
  dplyr::group_by(.data$i, .data$j_age_group) |>
  dplyr::group_modify(
    ~ .x |> 
      dplyr::mutate(
        "j" = sample(
          x = which(
            stringr::str_starts(population, .y$j_age_group) & seq_along(population) != .y$i
          ), 
          size = dplyr::n()
        )
      )
  ) |>
  dplyr::mutate( # Make lower
    "i_u" = pmax(.data$i, .data$j),
    "j_u" = pmin(.data$i, .data$j)
  ) |>
  dplyr::group_by(.data$i_u, .data$j_u) |>
  dplyr::summarise("n" = dplyr::n(), .groups = "drop")

# Convert to adjacency matrix with 2-way interactions
adjacency_sw <- as.matrix(
  Matrix::sparseMatrix(
    i = one_way_links$i_u, 
    j = one_way_links$j_u, 
    x = one_way_links$n,
    dims = c(length(population), length(population)),
    dimnames = list(population, population)
  )
)

#adjacency_sw <- floor(adjacency_sw / 2 + runif(length(population)^2)) # Prevent double counting

adjacency_sw <- (adjacency_sw + t(adjacency_sw)) / 2

print(adjacency_sw)
```

```{r, include = FALSE}
adjacency_graph <- function(adjacency) {
  graph <- igraph::graph_from_adjacency_matrix(
    adjmatrix = adjacency,
    mode = "undirected",
    diag = FALSE
  )

  region <- stringr::str_extract(population, r"{\w+$}")
  age <- stringr::str_extract(population, r"{^\w+}")

  layout <- cbind(
    x = as.numeric(factor(region)),
    y = as.numeric(factor(age))
  )

  # Add some jitter so vertices in the same region/age group do not overlap
  layout <- layout + matrix(
    stats::runif(length(layout), -0.3, 0.3),
    ncol = 2
  )

  plot(
    graph,
    layout = layout,
    vertex.color = factor(region),
    vertex.shape = "circle",
    vertex.label = population,
    vertex.size = stats::setNames(
      seq(
        from = 30,
        by = 5,
        along.with = unique(age)
      ),
      unique(age)
    )[age]
  )
}
```


```{r, eval = rlang::is_installed("igraph")}
adjacency_graph(adjacency_sw)
```

From this full adjacency graph, we can project the contacts ("T" domain) into contact matrices in different stratifications.

```{r}
contact_matrix <- function(adjacency, population, groups) {

  p <- purrr::map(
    unique(groups),
    ~ as.numeric(groups == .x)
  ) |>
    purrr::reduce(
      rbind, 
      .init = matrix(0, nrow = 0, ncol = length(population))
    )

  # Count population by group
  n_groups <- lengths(split(groups, factor(groups)))[unique(groups)]

  colnames(p) <- population
  rownames(p) <- paste0(names(n_groups), " (", n_groups, ")")

  return(p %*% adjacency %*% t(p))
}
```

Number of contacts stratified by age group and region
```{r}
T_age_region <- contact_matrix(
  adjacency = adjacency_sw,
  population = population,
  groups = population # No aggregation
)

print(T_age_region)
cat("Total contacts: ", sum(T_age_region))
```

Stratify only by region
```{r}
T_region <- contact_matrix(
  adjacency = adjacency_sw,
  population = population,
  groups = stringr::str_replace(population, r"{[A-Z]}", "*")
)

print(T_region)
cat("Total contacts: ", sum(T_region))
```

Stratify only by age group (and compare with input used to generate the adjacency graph)
```{r}
T_age <- contact_matrix(
  adjacency = adjacency_sw,
  population = population,
  groups = stringr::str_replace(population, r"{[a-z]}", "*")
)

print(T_age)
cat("Total contacts: ", sum(T_age))

cat("\n\nReference:\n")
print(t_matrix)
```


Merging age groups A and B
```{r}
T_merged_age_region <- contact_matrix(
  adjacency = adjacency_sw,
  population = population,
  groups = stringr::str_replace(population, r"{^A|B}", "AB")
)

print(T_merged_age_region)
cat("Total contacts: ", sum(T_merged_age_region))
```

Lets see if we can achieve these same contacts without generating the network.

```{r}
mixing_sw <- matrix(
  1 / length(regions),
  nrow = length(regions),
  ncol = length(regions),
  dimnames = list(regions, regions)
)
print(mixing_sw)
```


```{r}
c_matrix_full_sw <- kronecker(mixing_sw, c_matrix)
dimnames(c_matrix_full_sw) <- list(unique(population), unique(population))
print(round(c_matrix_full_sw, digits = 1))
```

```{r}
N <- lengths(split(population, factor(population))[unique(population)])
t_matrix_full_sw <- c_matrix_full_sw * outer(N, N)
print(round(t_matrix_full_sw, digits = 1))
```

```{r}
T_age_indirect <- contact_matrix(
  t_matrix_full_sw,
  population = unique(population),
  groups = stringr::str_replace(unique(population), r"{[a-z]}", "*")
)

print(round(T_age_indirect, digits = 1))
cat("Total contacts: ", round(sum(T_age_indirect), digits = 1))

cat("\n\nReference:\n")
print(t_matrix)
```

```{r}
N <- lengths(split(population, factor(population))[unique(population)])
print(N)
print(max(eigen(outer(N, N))$values))

N_age <- lengths(
  split(
    stringr::str_replace(population, r"{[a-z]}", "*"), 
    factor(stringr::str_replace(population, r"{[a-z]}", "*"))
  )
)
print(N_age)
print(max(eigen(outer(N_age, N_age))$values))

print(max(eigen(outer(N_age, N_age))$values) / max(eigen(outer(N, N))$values))
```

## A silo example
Consider a complementary network where a population of individuals across different age groups and
spatial locations are well-mixed only within their spatial region (a extremely "siloed" network).

We can draw a realisation of such a network from our contact matrices
```{r}
one_way_links <- tidyr::expand_grid(
  i_age_group = age_groups,
  j_age_group = age_groups
) |>
  dplyr::mutate(
    "n_contacts" = as.vector(t_matrix)
  ) |>
  tidyr::uncount(.data$n_contacts) |>
  dplyr::select(!"n_contacts") |>
  dplyr::group_by(.data$i_age_group) |>
  dplyr::group_modify(
    ~ {
      i_sampled <- .x |>
        dplyr::mutate(
          "i_age_group" = .y$i_age_group,
          "i" = sample(
            x = which(
              stringr::str_detect(population, .y$i_age_group)
            ), 
            size = dplyr::n()
          ),
          "i_lab" = population[.data$i],
          "i_region" = stringr::str_extract(.data$i_lab, "[a-z]")
        )
      
      # Determine reachable connections
      i_sampled <- i_sampled |> 
        dplyr::mutate(
          "any_j_reachable" = purrr::pmap_lgl(
            dplyr::select(i_sampled, "j_age_group", "i_region", "i"),
            \(j_age_group, i_region, i) {
              any(
                stringr::str_detect(population, stringr::str_replace(j_age_group, r"{\*}", i_region)) & 
                  seq_along(population) != i
              )
            }
          )
        )
      
      # Determine reachable connections
      i_accepted <- i_sampled |>
        dplyr::filter(.data$any_j_reachable)
      
      # Re-sample unreachable from reachable
      i_resampled <- i_sampled |>
        dplyr::filter(!.data$any_j_reachable) |>
        dplyr::count(.data$i_age_group, .data$j_age_group) |>
        purrr::pmap(
          \(i_age_group, j_age_group, n) {
            i_accepted |>
              dplyr::filter(.data$i_age_group == {{ i_age_group }}, .data$j_age_group == {{ j_age_group }}) |>
              dplyr::slice_sample(n = {{ n }})
          }
        ) |>
        purrr::list_rbind()
        
      return(dplyr::select(rbind(i_accepted, i_resampled), !"i_age_group"))
    }
  ) |> 
  dplyr::group_by(.data$i, .data$j_age_group, .data$i_region) |>
  dplyr::group_modify(
    ~ .x |> 
      dplyr::mutate(
        "j" = sample(
          x = which(
            stringr::str_detect(population, stringr::str_replace(.y$j_age_group, r"{\*}", .y$i_region)) & seq_along(population) != .y$i
          ), 
          size = dplyr::n()
        ),
        "j_lab" = population[.data$j],
        "j_region" = stringr::str_extract(.data$j_lab, "[a-z]")
      )
  ) |>
  dplyr::mutate( # Make lower
    "i_u" = pmax(.data$i, .data$j),
    "j_u" = pmin(.data$i, .data$j)
  ) |>
  dplyr::group_by(.data$i_u, .data$j_u) |>
  dplyr::summarise("n" = dplyr::n(), .groups = "drop")

# Convert to adjacency matrix with 2-way interactions
adjacency_silo <- as.matrix(
  Matrix::sparseMatrix(
    i = one_way_links$i_u, 
    j = one_way_links$j_u, 
    x = one_way_links$n,
    dims = c(length(population), length(population)),
    dimnames = list(population, population)
  )
)

#adjacency_silo <- floor(adjacency_silo / 2 + runif(length(population)^2)) # Prevent double counting

adjacency_silo <- (adjacency_silo + t(adjacency_silo)) / 2

print(adjacency_silo)
```

```{r, eval = rlang::is_installed("igraph")}
adjacency_graph(adjacency_silo)
```


Number of contacts stratified by age group and region
```{r}
T_age_region <- contact_matrix(
  adjacency = adjacency_silo,
  population = population,
  groups = population # No aggregation
)

print(T_age_region)
cat("Total contacts: ", sum(T_age_region))
```

Stratify only by region
```{r}
T_region <- contact_matrix(
  adjacency = adjacency_silo,
  population = population,
  groups = stringr::str_replace(population, r"{[A-Z]}", "*")
)

print(T_region)
cat("Total contacts: ", sum(T_region))
```

Stratify only by age group
```{r}
T_age <- contact_matrix(
  adjacency = adjacency_silo,
  population = population,
  groups = stringr::str_replace(population, r"{[a-z]}", "*")
)

print(T_age)
cat("Total contacts: ", sum(T_age))

cat("\n\nReference:\n")
print(t_matrix)
```


Merging age groups A and B
```{r}
T_merged_age_region <- contact_matrix(
  adjacency = adjacency_silo,
  population = population,
  groups = stringr::str_replace(population, r"{^A|B}", "AB")
)

print(T_merged_age_region)
cat("Total contacts: ", sum(T_merged_age_region))
```

Again, lets see if we can achieve these same contacts without generating the network.

```{r}
mixing_silo <- diag(
  1,
  nrow = length(regions),
  ncol = length(regions)
)
dimnames(mixing_silo) <- list(regions, regions)
print(mixing_silo)
```


```{r}
c_matrix_full_silo <- kronecker(mixing_silo, c_matrix)
dimnames(c_matrix_full_silo) <- list(unique(population), unique(population))
print(round(c_matrix_full_silo, digits = 1))
```

```{r}
N <- lengths(split(population, factor(population))[unique(population)])
t_matrix_full_silo <- c_matrix_full_silo * outer(N, N)
print(round(t_matrix_full_silo, digits = 1))
```

```{r}
T_age_indirect <- contact_matrix(
  t_matrix_full_silo,
  population = unique(population),
  groups = stringr::str_replace(unique(population), r"{[a-z]}", "*")
)

print(round(T_age_indirect, digits = 1))
cat("Total contacts: ", round(sum(T_age_indirect), digits = 1))

cat("\n\nReference:\n")
print(t_matrix)
```

```{r}
# Population in most stratified groups
N <- lengths(split(population, factor(population))[unique(population)])
print(N)
print(max(eigen(outer(N, N))$values))

# Population stratified by age
N_age <- lengths(
  split(
    stringr::str_replace(population, r"{[a-z]}", "*"), 
    factor(stringr::str_replace(population, r"{[a-z]}", "*"))
  )
)
print(N_age)
print(max(eigen(outer(N_age, N_age))$values))

# Scaling factor difference
s <- max(eigen(outer(N_age, N_age))$values) / max(eigen(outer(N, N))$values)
print(s)

print(round(s * T_age_indirect, digits = 1))
cat("Total contacts: ", round(sum(T_age_indirect), digits = 1))
```

```{r}

#adjacency_silo
idx <- split(
  seq_along(population), 
  stringr::str_extract(population, "[a-z]")
)[unique(stringr::str_extract(population, "[a-z]"))]

T_matrices <- purrr::map(
  idx,
  ~ adjacency_silo[., .]
) |>
  purrr::map(
    ~ contact_matrix(
      .,
      population = colnames(.),
      groups = stringr::str_replace(colnames(.), r"{[a-z]}", "*")
    )
  )

print(T_matrices)

C_matrices <- T_matrices |>
  purrr::map(
    ~ {
      N <- as.numeric(stringr::str_extract(colnames(.), r"{\d+}"))
      
      . / outer(N, N)
    }
  )

print(C_matrices)


```


## Mapping between examples

Notice that both examples build on the idea of "one contact" between individuals
and an idea of regional connectedness (small-world vs silo).

These examples are the extremes of what we may expect the real world to be like,
where individuals are likely to have many local contacts and fewer far-reaching
contacts. The spatial resolution that we impose on the model should not dictate
the number of contacts and we therefore need to correct for the differences in
the number of contacts seen in the toy-examples.

In `vignette("diseasy-regions")` we cover the idea of regional "connectedness"
(here "adjacency") and also derive the "neutral" connectedness across regions
which leaves the equations invariant to spatial stratification.

Here, we derive that movement probabilities of $\phi_{x,y} = 1 / \sqrt{n}$ for
an $n$-region system will be "neutral". The resulting "infection-flow" adjacency
from these movement probabilities are then $\Theta_{x,y} = 1  /n$.

Notice that the "adjacency" in our examples can be expressed in the
following decompositions:

Small-world example:

$$
\textrm{adjacency} = \begin{bmatrix}
  \overline{\overline{C}} & \overline{\overline{C}} \\
  \overline{\overline{C}} & \overline{\overline{C}}
\end{bmatrix} \cdot
\begin{bmatrix}
  \overline{\overline{1}} & \overline{\overline{1}} \\
  \overline{\overline{1}} & \overline{\overline{1}}
\end{bmatrix}
$$

Silo example:

$$
\textrm{adjacency} = \begin{bmatrix}
  \overline{\overline{C}} & \overline{\overline{C}} \\
  \overline{\overline{C}} & \overline{\overline{C}}
\end{bmatrix} \cdot
\begin{bmatrix}
  \overline{\overline{1}} & \overline{\overline{0}} \\
  \overline{\overline{0}} & \overline{\overline{1}}
\end{bmatrix}
$$

Where $\overline{\overline{C}}$ and $\overline{\overline{1}}$ is this example are
square unit matrices with dimensions equal to the number of age groups.
And $\overline{\overline{0}} = 0 \cdot \overline{\overline{1}}$

One way of summarising the effect of this adjacency on the SEIR models is to
consider maximum eigenvalue of the matrix which defines an overall scaling of
the transformation.

If we here compute these scaling factors
```{r}
print(max(eigen(adjacency_sw)$values))
print(max(eigen(adjacency_silo)$values))
```

... we see that the well-mixed small world example is exactly different in scaling
by the factor of $1/2$ expected from the considerations in neutral example.

If we have an intermediate adjacency matrix with some mixing across regions but
predominant mixing within regions we may get adjacencies such as:

```{r}
adjacency_mix <- outer(
  X = substr(population, 3, 3),
  Y = substr(population, 3, 3),
  FUN = \(x, y) ifelse(x == y, 0.8, 0.2)
)

colnames(adjacency_mix) <- rownames(adjacency_mix) <- population

print(adjacency_mix)
print(max(eigen(adjacency_mix)$values))
```
which for these parameters matches the scaling factor of the silo example.

If we instead consider only the regional mixing matrices of the above
adjacency matrices, we can isolate the effect of these mixing matrices alone.

```{r}
regional_mixing_sw   <- matrix(
  c(1,   1,   1,   1  ), nrow = 2, dimnames = list(c("x", "y"), c("x", "y"))
)
regional_mixing_silo <- matrix(
  c(1,   0,   0,   1  ), nrow = 2, dimnames = list(c("x", "y"), c("x", "y"))
)
regional_mixing_mix  <- matrix(
  c(0.8, 0.2, 0.2, 0.8), nrow = 2, dimnames = list(c("x", "y"), c("x", "y"))
)

print(regional_mixing_sw)
print(max(eigen(regional_mixing_sw)$values))

print(regional_mixing_silo)
print(max(eigen(regional_mixing_silo)$values))

print(regional_mixing_mix)
print(max(eigen(regional_mixing_mix)$values))
```
with the remaining scaling being due to the size of the population.

Notably, these results suggest that given a specific contact matrix derived
from survey data, there may be some regional length scale that these results
apply to and interpreting these contacts matrices at a different length scale
will introduce a different scaling constant. However, the most important aspect
of the contact matrices is the pattern of contacts across age groups and
therefore it is likely prudent to normalise both the regional mixing matrices
and the contact matrices to a unit scale and let the overall scaling factor
describe the strength of the interactions.


```{r cleanup, include = FALSE}
rm(
  regions,
  activity,
  population
)
gc()
```
