---
title: "SEIR: Implementation"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction
This vignette is a companion to the implementation of the SEIR ode model `?DiseasyModelOdeSeir`.

As the name suggests, this model is a SEIR-like ordinary differential equation (ODE) model.
Furthermore, the model is designed as flexible as possible to allow for a wide range of different model configurations.
Meaning that the number of consecutive exposed, infected, and recovered states can be set arbitrarily,
as well as the number of demographic groups (e.g. age groups or sex) and the number of variants.

Due to the flexibility of the model, the implementation is somewhat complex, which is why we include this vignette
outlining the mathematical implementation of the model and connect it to the code.

# Opening consideration
The design of the SEIR model is meant to be "standard" in the sense that it, in general, follows the structure and
design of SEIR models in the community.

For clarity, we here describe the considerations needed to design the model.

## Indexes and accents
To begin, here we define a fixed meaning to the different indices and accents used in the vignette:

| Symbol  | Description                |
|---------|----------------------------|
| $K,k$   | Exposed states             |
| $L,l$   | Infected states            |
| $M,m$   | Recovered states           |
| $D,p,q$ | Demographic groups         |
| $i,j$   | Age-group indicies         |
| $x,y$   | Regional indicies          |
| $V,a,b$ | Variants                   |
| $^\sim$ | Non-normalised quantity    |

Where upper case letters denote the number of such elements and lower case letters denote a specific element.

## The model structure
The SEIR model we are building is a compartmental model with an (near) arbitrary number of exposed, infected,
and recovered compartments. The model also supports multiple demographic groups and multiple variants.

We divide the model into different "tracks" with each track implementing the disease progression for a group of
individuals.
That is, if we have $D$ demographic groups and $V$ variants, we have $D \cdot V$ tracks which each start with the first
exposed state (or the first infected state if no exposed states are included in the model).
As the infection progresses within each individual, they are moved down the track through infection and eventually
into recovery.

Schematically, the model can be represented as in the figure below:

```{r SEIR-structure, echo = FALSE, fig.align = "center", fig.cap = "SEIR model overview for a two variant model", out.width = "120%"}
knitr::include_graphics(
  devtools::package_file(
    "vignettes/articles/article_data/SEIR-implementation-model.png"
  )
)
```

This figure shows the disease progression for two variant tracks for a single demographic group.
The arrows in the figure indicates the infections in the model within these tracks.

## State vector

Each "track" of infection is placed in a sequence in the state vector, $\overline{\psi}$, and increments first in demographic
groups then in variants.

Finally, the susceptible states are placed at the end of the state vector[^1].

[^1]: If the user has configured custom model outputs (see the "DiseasyModelOde" article) each such output needs a state
(here called `C_\cdot` state) which is placed after the susceptible states in the state vector.

$$
\overline{\psi} = [\underbrace{\begin{matrix} E_{1,1}^1\, \cdots \, E_{1,1}^K \;\;\; I_{1,1}^1 \, \cdots \, I_{1,1}^L \;\;\; R_{1,1}^1 \, \cdots \, R_{1,1}^M \end{matrix}}_{\text{Demographic group 1, Variant 1}}
$$
$$
\underbrace{\begin{matrix} E_{2,1}^1 \, \cdots \, E_{2,1}^K \;\;\; I_{2,1}^1 \, \cdots \, I_{2,1}^L \;\;\; R_{2,1}^1 \, \cdots \, R_{2,1}^M \end{matrix}}_{\text{Demographic group 2, Variant 1}}
\; \cdots
$$
$$
\underbrace{\begin{matrix} E_{D,V}^1 \, \cdots \, E_{D,V}^K \;\;\; I_{D,V}^1 \, \cdots \, I_{D,V}^L \;\;\; R_{D,V}^1 \, \cdots \, R_{D,V}^M \end{matrix}}_{\text{Demographic group D, Variant V}}
\;\; S_1 \, \cdots \, S_D]
$$
NOTE: State vector uses normalised (unit) populations.

Notice that the state vector elements are indexed by state-type, state-index, demography-index and variant-index:

```{r SERI-index, echo = FALSE, fig.align = "center", fig.cap = "Index guide for the state vector"}
knitr::include_graphics(
  devtools::package_file(
    "vignettes/articles/article_data/SEIR-implementation-index.png"
  )
)
```

## Contact matrix
If we define a risk of infection per contact, $P$, the differential equation governing the infectious in the simple SIR
model can then be expressed as:
$$
\dot{\tilde{I}} =
\left<\textrm{risk of infection per contact}\right> \times
$$
$$
\left<\textrm{number of contacts from infected individuals}\right> \times \left<\textrm{probability of contact being susceptible}\right>
$$
$$
 - \left<\textrm{clearing of infection}\right>
$$

To compute the number of infectious contacts we need the number of contacts between infected individuals of any demographic
group and individuals in the $p$'th demographic group. If $t_{p,q} = t_{q,p}$ is the (symmetric) number of contacts between
demographic group $p$ and demographic group $q$, then $P \tilde{I}_q t_{p,q} / \tilde{N}_q$ is the number of infectious contacts from
$q$ to $p$
(infected individual of demographic group $q$ times mean number of contacts to demographic group $p$ with each contact carrying a
risk of infection $P$).

The probability that a contact is susceptible is simply its proportion relative to the demographic group leading to:

$$
\dot{\tilde{I}}_p = P\left(\tilde{I}_1 t_{p,1} / \tilde{N}_1 + \tilde{I}_2 t_{p,2} / \tilde{N}_2 + \dots\right) \cdot
\frac{\tilde{S}_p}{\tilde{N}_p} - r_I \tilde{I}_p
$$

Collecting $N$ terms:
$$
= P\left(\tilde{I}_1 t_{p,1} / (\tilde{N}_1 \tilde{N}_p) + \tilde{I}_2 t_{p,2} / (\tilde{N}_2 \tilde{N}_p) + \dots\right)
\cdot \tilde{S}_p - r_I \tilde{I}_p \Rightarrow
$$

Then, dividing through by the total population size $N$ yields:
$$
\frac{\dot{\tilde{I}}_p}{\tilde{N}} = P\left(\tilde{I}_1 t_{p,1} / (\tilde{N}_1 \tilde{N}_p) + \tilde{I}_2 t_{p,2} / (\tilde{N}_2 \tilde{N}_p) + \dots\right)
\cdot \frac{\tilde{S}_p}{\tilde{N}} - r_I \frac{\tilde{I}_p}{\tilde{N}} \Rightarrow
$$

Then expanding by $N / N$ and expressing in terms of $c_{p,q} = \frac{t_{p,q}}{\tilde{N}_p\tilde{N}_q}$.
$$
\frac{\dot{\tilde{I}}_p}{\tilde{N}} = P\frac{\tilde{N}}{\tilde{N}}\left(\tilde{I}_1 c_{p,1} + \tilde{I}_2 c_{p,2} + \dots\right) \cdot
\frac{\tilde{S}_p}{\tilde{N}}
- r_I \frac{\tilde{I}_p}{\tilde{N}} \Rightarrow
$$

and finally expressing in terms of normalised quantities
$$
\dot{I}_p = P\tilde{N}\left(I_1 c_{p,1} + I_2 c_{p,2} + \dots\right) \cdot
S_p - r_I I_p
$$

Notice that we obtain terms of the form:
$$
\tilde{N}c_{p,q} = \frac{\tilde{N}}{\tilde{N}_q} \frac{t_{p,q}}{\tilde{N}_p} = \frac{\tilde{N}}{\tilde{N}_q} m_{p,q}
$$


## Regional aggregation of contact matrices
The source data for the contact matrices we use in `diseasy` is typically collections of survey data where
participants have been instructed to count contacts with other individuals along with the approximate age of
these individuals. These survey data are then corrected for differences in demography of the participants
and the group they are meant to represent (e.g. a given country) for form a "full" contact matrix for the
target population.

To gain some intuition on the how these contact matrices can be interpreted and manipulated we will
begin by considering some toy-examples with a toy population of age groups ("A", "B", "C") and spatial
regions ("x", "y").

```{r}
population <- c(
  rep("A/x", 1), rep("B/x", 2), rep("C/x", 3),
  rep("A/y", 3), rep("B/y", 2), rep("C/y", 1)
)
```

### A small-world example
Consider a the fully-connected case where a population of individuals across different age groups and
spatial locations are all connected by one contact (a extremely "small-world" network).

```{r}
adjacency_sw <- matrix(1, nrow = length(population), ncol = length(population))
colnames(adjacency_sw) <- rownames(adjacency_sw) <- population
print(adjacency_sw)
```

```{r, include = FALSE}
adjacency_graph <- function(adjacency) {
  graph <- igraph::graph_from_adjacency_matrix(
    adjmatrix = adjacency,
    mode = "undirected",
    diag = FALSE
  )

  region <- stringr::str_extract(population, r"{\w+$}")
  age <- stringr::str_extract(population, r"{^\w+}")

  layout <- cbind(
    x = as.numeric(factor(region)),
    y = as.numeric(factor(age))
  )

  # Add some jitter so vertices in the same region/age group do not overlap
  layout <- layout + matrix(
    stats::runif(length(layout), -0.3, 0.3),
    ncol = 2
  )

  plot(
    graph,
    layout = layout,
    vertex.color = factor(region),
    vertex.shape = "circle",
    vertex.label = population,
    vertex.size = stats::setNames(
      seq(
        from = 30,
        by = 5,
        along.with = unique(age)
      ),
      unique(age)
    )[age]
  )
}
```

```{r, eval = rlang::is_installed("igraph")}
adjacency_graph(adjacency_sw)
```


```{r}
contact_matrix <- function(adjacency, population, groups) {

  p <- lengths(
    split(
      population,
      factor(groups, levels = unique(groups))
    )
  )

  p <- purrr::map(
    unique(groups),
    ~ as.numeric(groups == .x)
  ) |>
    purrr::reduce(rbind) |>
    as.matrix()

  colnames(p) <- population
  rownames(p) <- unique(groups)

  return(p %*% adjacency %*% t(p) - diag(rowSums(p)))
}
```

Number of contacts stratified by age group and region
```{r}
T_age_region <- contact_matrix(
  adjacency = adjacency_sw,
  population = population,
  groups = population # No aggregation
)

print(T_age_region)
print(sum(T_age_region))
```

Stratify only by region
```{r}
T_region <- contact_matrix(
  adjacency = adjacency_sw,
  population = population,
  groups = stringr::str_replace(population, r"{[A-Z]}", "*")
)

print(T_region)
print(sum(T_region))
```

Stratify only by age group
```{r}
T_age <- contact_matrix(
  adjacency = adjacency_sw,
  population = population,
  groups = stringr::str_replace(population, r"{[a-z]}", "*")
)

print(T_age)
print(sum(T_age))
```


Merging age groups A and B
```{r}
T_merged_age_region <- contact_matrix(
  adjacency = adjacency_sw,
  population = population,
  groups = stringr::str_replace(population, r"{^A|B}", "AB")
)

print(T_merged_age_region)
print(sum(T_merged_age_region))
```


### A silo example
Consider a complementary network where a population of individuals across different age groups and
spatial locations are all connected by one contact only within their spatial region (a extremely "siloed" network).

```{r}
adjacency_silo <- outer(
  X = substr(population, 3, 3),
  Y = substr(population, 3, 3),
  FUN = \(x, y) as.numeric(x == y)
)
colnames(adjacency_silo) <- population
rownames(adjacency_silo) <- population
print(adjacency_silo)
```

```{r, eval = rlang::is_installed("igraph")}
adjacency_graph(adjacency_silo)
```


Number of contacts stratified by age group and region
```{r}
T_age_region <- contact_matrix(
  adjacency = adjacency_silo,
  population = population,
  groups = population # No aggregation
)

print(T_age_region)
print(sum(T_age_region))
```

Stratify only by region
```{r}
T_region <- contact_matrix(
  adjacency = adjacency_silo,
  population = population,
  groups = stringr::str_replace(population, r"{[A-Z]}", "*")
)

print(T_region)
print(sum(T_region))
```

Stratify only by age group
```{r}
T_age <- contact_matrix(
  adjacency = adjacency_silo,
  population = population,
  groups = stringr::str_replace(population, r"{[a-z]}", "*")
)

print(T_age)
print(sum(T_age))
```


Merging age groups A and B
```{r}
T_merged_age_region <- contact_matrix(
  adjacency = adjacency_silo,
  population = population,
  groups = stringr::str_replace(population, r"{^A|B}", "AB")
)

print(T_merged_age_region)
print(sum(T_merged_age_region))
```

## Mapping between examples
Notice that both examples build on the idea of "one contact" between individuals
and an idea of regional connectedness (small-world vs silo).

These examples are the extremes of what we may expect the real world to be like,
where individuals are likely to have many local contacts and fewer far-reaching
contacts. The spatial resolution that we impose on the model should not dictate
the number of contacts and we therefore need to correct for the differences in
the number of contacts seen in the toy-examples.

In `vignette("diseasy-regions")` we cover the idea of regional "connectedness"
(here "adjacency") and also derive the "neutral" connectedness across regions
which leaves the equations invariant to spatial stratification.

Here, we derive that movement probabilities of $\phi_{x,y} = 1 / \sqrt{n}$ for
an $n$-region system will be "neutral". The resulting "infection-flow" adjacency
from these movement probabilities are then $\Theta_{x,y} = 1  /n$.

Notice that the "adjacency" in our examples can be expressed in the
following decompositions:

Small-world example:

$$
\textrm{adjacency} = \begin{bmatrix}
  \overline{\overline{C}} & \overline{\overline{C}} \\
  \overline{\overline{C}} & \overline{\overline{C}}
\end{bmatrix} \cdot
\begin{bmatrix}
  \overline{\overline{1}} & \overline{\overline{1}} \\
  \overline{\overline{1}} & \overline{\overline{1}}
\end{bmatrix}
$$

Silo example:

$$
\textrm{adjacency} = \begin{bmatrix}
  \overline{\overline{C}} & \overline{\overline{C}} \\
  \overline{\overline{C}} & \overline{\overline{C}}
\end{bmatrix} \cdot
\begin{bmatrix}
  \overline{\overline{1}} & \overline{\overline{0}} \\
  \overline{\overline{0}} & \overline{\overline{1}}
\end{bmatrix}
$$

Where $\overline{\overline{C}}$ and $\overline{\overline{1}}$ is this example are
square unit matrices with dimensions equal to the number of age groups.
And $\overline{\overline{0}} = 0 \cdot \overline{\overline{1}}$

One way of summarising the effect of this adjacency on the SEIR models is to
consider maximum eigenvalue of the matrix which defines an overall scaling of
the transformation.

If we here compute these scaling factors
```{r}
print(max(eigen(adjacency_sw)$values))
print(max(eigen(adjacency_silo)$values))
```

... we see that the well-mixed small world example is exactly different in scaling
by the factor of $1/2$ expected from the considerations in neutral example.

If we have an intermediate adjacency matrix with some mixing across regions but
predominant mixing within regions we may get adjacencies such as:

```{r}
adjacency_mix <- outer(
  X = substr(population, 3, 3),
  Y = substr(population, 3, 3),
  FUN = \(x, y) ifelse(x == y, 0.8, 0.2)
)

colnames(adjacency_mix) <- rownames(adjacency_mix) <- population

print(adjacency_mix)
print(max(eigen(adjacency_mix)$values))
```
which for these parameters matches the scaling factor of the silo example.

If we instead consider only the regional mixing matrices of the above
adjacency matrices, we can isolate the effect of these mixing matrices alone.

```{r}
regional_mixing_sw   <- matrix(
  c(1,   1,   1,   1  ), nrow = 2, dimnames = list(c("x", "y"), c("x", "y"))
)
regional_mixing_silo <- matrix(
  c(1,   0,   0,   1  ), nrow = 2, dimnames = list(c("x", "y"), c("x", "y"))
)
regional_mixing_mix  <- matrix(
  c(0.8, 0.2, 0.2, 0.8), nrow = 2, dimnames = list(c("x", "y"), c("x", "y"))
)

print(regional_mixing_sw)
print(max(eigen(regional_mixing_sw)$values))

print(regional_mixing_silo)
print(max(eigen(regional_mixing_silo)$values))

print(regional_mixing_mix)
print(max(eigen(regional_mixing_mix)$values))
```
with the remaining scaling being due to the size of the population.

Notably, these results suggest that given a specific contact matrix derived
from survey data, there may be some regional length scale that these results
apply to and interpreting these contacts matrices at a different length scale
will introduce a different scaling constant. However, the most important aspect
of the contact matrices is the pattern of contacts across age groups and
therefore it is likely prudent to normalise both the regional mixing matrices
and the contact matrices to a unit scale and let the overall scaling factor
describe the strength of the interactions.


# The implementation of the right-hand side function
With the opening considerations in mind, we can now describe the implementation of the model.
Below, we now link the implementation with the underlying mathematical descriptions of the model.

The `?DiseasyModelOdeSeir` model implements the `DiseasyModelOdeSeir$rhs` method which computes the
rates needed for the ODE solver. The evaluation of this right-hand side function uses a few intermediate
variables which are described below.

## The `infected` matrix
In the computation of the right-hand side of the ODE, we have a helper matrix $\overline{\overline{I}}$ stored
as `infected` in the implementation.

This $\overline{\overline{I}}$ matrix is a $D \times V$ matrix with elements

$$
\overline{\overline{I}} = \begin{bmatrix}
\sum_l I^l_{1,1} & \cdots & \sum_l I^l_{1,V} \\
\vdots & \ddots & \vdots \\
\sum_l I^l_{D,1} & \cdots & \sum_l I^l_{D,V}
\end{bmatrix}
$$

That is, the element $i_{a, v}$ contains the sum of all infected compartments for the track associated with
demographic group $a$ and variant $v$.


## The `infected_contact_rate` matrix

During the right-hand side computation, we also have a helper matrix `infected_contact_rate` which is
$\overline{\overline{C}}\; \overline{\overline{I}}$

$$
\overline{\overline{CI}} = \overline{\overline{C}}\; \overline{\overline{I}} = \begin{bmatrix}
\sum_p c_{1\leftarrow p} \sum_l I^l_{p,1} & \cdots & \sum_p c_{1\leftarrow p} \sum_l I^l_{p,V} \\
\vdots & \ddots & \vdots \\
\sum_p c_{D\leftarrow p} \sum_l I^l_{p,1} & \cdots & \sum_p c_{D\leftarrow p} \sum_l I^l_{p,V}
\end{bmatrix}
$$

That is, the element $(\overline{\overline{CI}})_{p, a}$ contains the rate of contact for demographic
group $p$ with persons infected with variant $a$ across all demographic groups.

## The `infection_rate` matrix

Continuing the computation of the right-hand side, we have the `infection_rate` matrix ($\overline{\overline{T}}$)
which is the `infected_contact_rate` ($\overline{\overline{CI}}$) adjusted for additional risk factors.

- $\sigma(t)$: Multiplicative effect of season
- $\epsilon_a$: Relative infectiousness of variant $a$
- $\Gamma$: Overall infection rate scaling factor

$$
\overline{\overline{T}} = \Gamma\;\sigma(t)\;\overline{\overline{\mathcal{E}}} \circ \overline{\overline{CI}}
$$

Where $\circ$ is the Hadamard product (element-wise multiplication), and $\overline{\overline{\mathcal{E}}}$ is the
$D \times V$ matrix:

$$
\overline{\overline{\mathcal{E}}} = \begin{bmatrix}
\epsilon_1 & \cdots & \epsilon_V \\
\vdots & \ddots & \vdots \\
\epsilon_1 & \cdots & \epsilon_V
\end{bmatrix}
$$

NOTE: In the code, `private$indexed_variant_infection_risk` is $\overline{\overline{\mathcal{E}}}$ in vector form.

## The `infection_matrix`
For the final computation of infections of the right-hand side, we construct the matrix `infection_matrix`
($\overline{\overline{f}}$).

Since only $R$ and $S$ compartments are susceptible to infection, we need only to consider the $RS$ states to
compute the new infections in the model.

The new `infection_matrix` therefore does not need to contain the $K$ exposed states and the $L$ infected states,
but only the $M$ recovered states and the $D$ susceptible states.

We are therefore interested in the reduced state vector $\overline{\psi}^{RS}$ whose elements $e^*$ are the $R$
and $S$ elements of the full state vector $\overline{\psi}$.


We now create an intermediate matrix $\overline{\overline{U}}$ by expanding the `infection_rate` matrix
($\overline{\overline{T}}$) to a $D \cdot (V \cdot M + 1) \times V$ matrix by repeating the rows of the
`infection_rate` matrix.
This repeating is done by matching the demographic group of the `infection_rate` matrix to the demographic group of the reduced state
vector.
Each element, $e$, in the state vector, $\psi$ has a corresponding demographic group which we can write as $p|e$.

The intermediate matrix $\overline{\overline{U}}$, has a row for all states in $R$ or $S$ compartments and column
for each variant. The form is as follows:

$$
\overline{\overline{U}} = \begin{bmatrix}
  t_{p|e^*_1, 1} & \cdots & t_{p|e^*_1, V} \\
  t_{p|e^*_2, 1} & \cdots & t_{p|e^*_2, V} \\
  \vdots & \vdots &\vdots \\
  t_{p|e^*_{D \cdot (V \cdot M + 1)}, 1} & \cdots & t_{p|e^*_{D \cdot (V \cdot M + 1)}, V} \\
\end{bmatrix}
$$

We now have a matrix, whose elements are the infection rate for each variant matched to the $R$ and $S$ elements
of the state vector.

In this space, we can now account for:

- The state specific infection risks (i.e. the reduced risk from being infected previously / vaccinated)
- The effect of cross-immunity between the variants

That is, we need to construct the matrix that accounts for the waning of immunity ($\gamma_m$) and the cross-immunity
factors $\chi_{a\leftarrow b}$ (Variant $b$ infecting a person with immunity for variant $a$).

All together, this is implemented via the immunity matrix (`immunity_matrix`) $\overline{\overline{\rho}}$ which is
$D \cdot (V \cdot M + 1) \times V$ matrix on the form:

$$
\overline{\overline{\rho}} = \begin{bmatrix}
  \begin{bmatrix}
    1 - \chi_{1\leftarrow 1} (1 - \overline{\gamma}) \\
    \vdots \\
    1 - \chi_{1\leftarrow 1} (1 - \overline{\gamma})
  \end{bmatrix}
  & \cdots &
  \begin{bmatrix}
    1 - \chi_{1\leftarrow V} (1 - \overline{\gamma}) \\
    \vdots \\
    1 - \chi_{1\leftarrow V} (1 - \overline{\gamma})
  \end{bmatrix} \\
  \vdots & \ddots & \vdots \\
  \begin{bmatrix}
    1 - \chi_{V\leftarrow 1} (1 - \overline{\gamma}) \\
    \vdots \\
    1 - \chi_{V\leftarrow 1} (1 - \overline{\gamma})
  \end{bmatrix}
  & \cdots &
  \begin{bmatrix}
    1 - \chi_{V\leftarrow V} (1 - \overline{\gamma}) \\
    \vdots \\
    1 - \chi_{V\leftarrow V} (1 - \overline{\gamma})
  \end{bmatrix} \\
  \overline{1}_D & \cdots & \overline{1}_D
\end{bmatrix}
$$

Where $\overline{1}_D$ is a ones vector of length $D$.

In $\overline{\overline{\rho}}$, the sub-matrices of the form:

$$
\begin{bmatrix}
  1 - \chi_{a\leftarrow b} (1 - \overline{\gamma}) \\
  \vdots \\
  1 - \chi_{a\leftarrow b} (1 - \overline{\gamma})
\end{bmatrix}
$$

are $D \cdot M \times 1$ matrices with repeated sub-matrices.
That is, $1 - \chi_{a\leftarrow b} (1 - \overline{\gamma})$ is a vector of length $M$ which is repeated
for each demographic group, to form the $D \cdot M \times 1$ matrix.

In total, the matrix accounts for the cross-variant $\chi_{a\leftarrow b}$ reductions of immunity ($1-\overline{\gamma}$).

Finally, we can combine everything to the flow matrix:

$$
\overline{\overline{f}} = \overline{\overline{\rho}} \odot
\begin{bmatrix}
  \vert &  & \vert \\
  \overline{\psi}^{RS} & \cdots & \overline{\psi}^{RS} \\
  \vert &  & \vert \\
\end{bmatrix} \odot
\overline{\overline{U}}
$$

Where $\begin{bmatrix}\overline{\psi}^{RS} & \cdots & \overline{\psi}^{RS}\end{bmatrix}$ is $V$ repeats of the
$R$ and $S$ elements of the state vector $\overline{\psi}$.

This matrix contains, for each $R$ and $S$ element in the state vector, a row consisting of per-variant infectious
contacts between the corresponding compartment and the infected population in the model. These elements account for
every risk modifying mechanism included in the model.

The row sums of this matrix is then equal to the loss from the compartment due to infectious contacts.
The sums by variant/demographic group combination can also be extracted from the matrix to form the in-flow of new infections
to the $E_1$ compartments.


# Generator matrix
Beyond implementing the right-hand side of the ODE, we also implement the generator matrix for the model.

The generator matrix, $\overline{\overline{G}}$, is a $D \cdot V \cdot (K+L) \times D \cdot V \cdot (K+L)$ matrix
containing the dynamics of the $E$ and $I$ states under the assumption of static $R$ and $S$ states.

Formally, the generator matrix is the linearisation:

$$
\frac{d\overline{\psi}^{EI}}{d t} = \overline{\overline{G}} \overline{\psi}^{EI}
$$

The matrix can be decomposed into two contributions:

- The transition matrix, $\overline{\overline{G}}_T$, which is a bi-diagonal matrix containing the flow rates between
  sequential compartments
- The transmission matrix, $\overline{\overline{G}}_\beta$ which contains the flows stemming from infections.

The transition matrix is block-diagonal and has the form (here for $K = 1$ and $L = 1$ and $D + V > 2)$:

$$
\overline{\overline{G}}_T =
\left[\begin{array}{cccc}
  \begin{bmatrix}
    -r_e & 0 \\
    r_e & -r_i
  \end{bmatrix} &
  \begin{matrix}
    0 & 0 \\
    0 & 0
  \end{matrix} &
  \cdots &
  \begin{matrix}
    0 & 0 \\
    0 & 0
  \end{matrix} \\
  \begin{matrix}
    0 & 0 \\
    0 & 0
  \end{matrix} &
  \begin{bmatrix}
    -r_e & 0 \\
    r_e & -r_i
  \end{bmatrix} &
  \cdots &
  \begin{matrix}
    0 & 0 \\
    0 & 0
  \end{matrix} \\
  \vdots & \vdots & \ddots & \vdots \\
  \begin{matrix}
    0 & 0 \\
    0 & 0
  \end{matrix} &
  \begin{matrix}
    0 & 0 \\
    0 & 0
  \end{matrix} &
  \cdots &
  \begin{bmatrix}
    -r_e & 0 \\
    r_e & -r_i
  \end{bmatrix}
\end{array}\right]
$$

Notice that the off-diagonal has zero-elements where the (reduced) state vector changes demographic groups
($I_p^L \rightarrow E_{p+1}^1$).

The form of the transmission matrix is more difficult to write up.

We lump risk modifying factors such as the effect of season $\sigma(t)$ and overall infection rate $\Gamma$ into
a single factor $\beta$.

Some examples:

### Two demographic groups, single variant

As long as we have only a single variant, cross immunity can be ignored and we can lump the effect of the balance
between $R$ to $S$ into $\beta$.


For $K = 1$ and $L = 1$, the transmission matrix is:

$$
\overline{\overline{G}}_\beta = \beta
\begin{bmatrix}
  0 & c_{1\leftarrow 1} & 0 & c_{1\leftarrow 2} \\
  0 & 0 & 0 & 0 \\
  0 & c_{2\leftarrow 1} & 0 & c_{2\leftarrow 2} \\
  0 & 0 & 0 & 0
\end{bmatrix}
$$

For $K = 1$ and $L = 2$, the transmission matrix is:

$$
\overline{\overline{G}}_\beta = \beta
\begin{bmatrix}
  0 & c_{1\leftarrow 1} & c_{1\leftarrow 1} & 0 & c_{1\leftarrow 2} & c_{1\leftarrow 2} \\
  0 & 0 & 0 & 0 & 0 & 0\\
  0 & 0 & 0 & 0 & 0 & 0\\
  0 & c_{2\leftarrow 1} & c_{2\leftarrow 1} & 0 & c_{2\leftarrow 2} & c_{2\leftarrow 2} \\
  0 & 0 & 0 & 0 & 0 & 0\\
  0 & 0 & 0 & 0 & 0 & 0
\end{bmatrix}
$$

### Single demographic group, double variant
With only one demographic group, the contact matrix is the scalar $c$.

With two variants, we need to also account for the relative infectivity of the variants $\epsilon_a$.

Then, for $K = 1$ and $L = 1$, the transmission matrix is:

$$
\overline{\overline{G}}_\beta = \beta\left[
\begin{array}{ccccc}
  0 & \epsilon_1\hat{\rho}_{1}\,c & | & 0 & 0 \\
  0 & 0 & | & 0 & 0 \\
  - & - & + & - & - \\
  0 & 0 & | & 0 & \epsilon_2\hat{\rho}_{2}\,c \\
  0 & 0 & | & 0 & 0
\end{array}
\right]
$$

Where the change between variants is demarcated visually by the vertical and horizontal lines.
We now have the additional factor $\hat{\rho}$ which explicitly accounts for the balance of $S$ and $R$ states and
the cross-immunity between the variants:

$$
\hat{\rho}_{a} = S + \sum_{b = 1}^V\sum_{m=1}^M \left(1 - \chi_{b \leftarrow a}(1 - \gamma_m)\right) R^m_b
$$

That is, we have the unmodified contribution from the $S$ states, and the contribution from all $R$ states while
accounting for waning of immunity ($\gamma_m$) and cross-immunity ($\chi_{a \leftarrow b}$).

This factor is analogous to a "cross-section" of infection encounter, i.e. a factor that accounts for the probability
that a contact between an infectious and non-infected person is an infectious contact.

Then, for $K = 1$ and $L = 2$, the transmission matrix is:

$$
\overline{\overline{G}}_\beta = \beta\left[
\begin{array}{ccccccc}
  0 & \epsilon_1\hat{\rho}_{1}\,c & \epsilon_1\hat{\rho}_{1}\,c & | & 0 & 0 & 0 \\
  0 & 0 & 0 & | & 0 & 0 & 0 \\
  0 & 0 & 0 & | & 0 & 0 & 0 \\
  - & - & - & + & - & - & - \\
  0 & 0 & 0 & | & 0 & \epsilon_2\hat{\rho}_{2}\,c & \epsilon_2\hat{\rho}_{2}\,c \\
  0 & 0 & 0 & | & 0 & 0 & 0 \\
  0 & 0 & 0 & | & 0 & 0 & 0 \\
\end{array}
\right]
$$

### Two demographic groups, double variant

When including demographic groups, the $\hat{\rho}$ factor becomes a $D \times V$ matrix of elements

$$
\hat{\rho}_{p, a} = S_p + \sum_{b = 1}^V\sum_{m=1}^M \left(1 - \chi_{b \leftarrow a}(1 - \gamma_m)\right) R^m_{p,b}
$$

For $K = 1$ and $L = 1$, the transmission matrix is:

$$
\overline{\overline{G}}_\beta = \beta\left[
\begin{array}{cccccccc}
  0 & \epsilon_1\hat{\rho}_{1,1}\,c_{1\leftarrow 1} & 0 & \epsilon_1\hat{\rho}_{1,1}\,c_{1\leftarrow 2} & | & 0 & 0 & 0 & 0 \\
  0 & 0 & 0 & 0 & | & 0 & 0 & 0 & 0\\
  0 & \epsilon_1\hat{\rho}_{2,1}\,c_{2\leftarrow 1} & 0 & \epsilon_1\hat{\rho}_{2,1}\,c_{2\leftarrow 2} & | & 0 & 0 & 0 & 0 \\
  0 & 0 & 0 & 0 & | & 0 & 0 & 0 & 0 \\
  - & - & - & - & + & - & - & - & - \\
  0 & 0 & 0 & 0 & | & 0 & \epsilon_2\hat{\rho}_{1,2}\,c_{1\leftarrow 1} & 0 & \epsilon_2\hat{\rho}_{1,2}\,c_{1\leftarrow 2} \\
  0 & 0 & 0 & 0 & | & 0 & 0 & 0 & 0 \\
  0 & 0 & 0 & 0 & | & 0 & \epsilon_2\hat{\rho}_{2,2}\,c_{2\leftarrow 1} & 0 & \epsilon_2\hat{\rho}_{2,2}\,c_{2\leftarrow 2} \\
  0 & 0 & 0 & 0 & | & 0 & 0 & 0 & 0
\end{array}
\right]
$$


# Notes on optimisations of the right-hand side function

It is important that the implementation is somewhat optimised for the simulations to run in a reasonable time frame.
To that end, we here show micro-benchmarks of different implementations of the same steps with the time consumption
as output, to highlight why a given implementation was chosen.

These benchmarks assume a simple SEIR model with only three demographic groups and one variant.

## "infected" sum computation
```{r, eval = rlang::is_installed("microbenchmark")}
## Step 1, determine the number of infected by demographic group and variant
i_state_indices <- list(2, 5, 8)
state_vector <- rep(0, 9)

# Benchmark possible approaches
microbenchmark::microbenchmark( # Microseconds
  "purrr::map_dbl" = purrr::map_dbl(
    i_state_indices,
    \(indices) sum(state_vector[indices])
  ),
  "sapply" = sapply(
    i_state_indices,
    \(indices) sum(state_vector[indices])
  ),
  "sapply - USE.NAMES" = sapply(
    i_state_indices,
    \(indices) sum(state_vector[indices]),
    USE.NAMES = FALSE
  ),
  "vapply" = vapply(
    i_state_indices,
    \(indices) sum(state_vector[indices]),
    FUN.VALUE = numeric(1),
    USE.NAMES = FALSE
  ),
  check = "equal", times = 1000L
)
```
Note, that the best performing function in this case is `vapply()`.
